---
title: "Severity Network Construction"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Identify pathways that are directly associated with COVID disease severity
```{r setup, include=FALSE}
data.folder <- "input.data"
output.folder <- "output/severity.network"
dir.create(output.folder,recursive = T)
library(knitr)
library(ggplot2)
library(pheatmap)
library(huge)
library(network)
library(ggnetwork)
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)
severity.color <- c("Critical-Alive"="#374E55FF","Critical-Deceased"="#DF8F44FF","Moderate-Alive"="#00A1D5FF" ,"Severe-Alive"="#B24745FF")
```

### Import Data
Use GSVA scores from the PC1_module_score_gsva_filtered_samples_genes file, which was created using edge genes of GSEA of PC1 coefficients.

Consider only a subset of pathways that are of interest and passed GSEA p value cutoff of 0.2
```{r import.data}
end.point.info <- readRDS(file.path(output.folder,"..","DSM","citeseq.patient.end.points.RDS"))
# selected gene sets from Fig 3 to analyse
selected.gene.sets <- read.csv(file.path(data.folder,"pathway_summary.csv"))

# GSEA stats
# gene sets that are sig. in GSEA (or in manual node list) and are of interest
gsea.res <- read.csv(file.path(data.folder,"gene.set.scores.and.gsea.csv"))
gsea.res$cell.type.pathway <- paste0(gsea.res$pathway,"_",gsea.res$celltype)
sig.gsea.res <- subset(gsea.res,PC1continuous_padj <= 0.2)
pathway.info <- unique(sig.gsea.res[,-c(11:15)])
pathway.info <- merge(selected.gene.sets,pathway.info,by.x="Primary_gene_sets",by.y="pathway")
rownames(pathway.info) <- pathway.info$cell.type.pathway

# gsva module scores
#module.score.type <- "continuousPC1_module_score_gsva_all_samples"
module.score.type <- "PC1_PC1_module_score_gsva_filtered_samples_genes"
gsva.scores <- readRDS(file.path(data.folder,paste0(module.score.type,".rds")))
gsva.scores$cell.type.pathway <- paste0(gsva.scores$pathway,"_",gsva.scores$celltype)
# keep only those that are sig. in GSEA
sig.gsva.scores <- subset(gsva.scores,cell.type.pathway %in% pathway.info$cell.type.pathway)
sig.gsva.scores[,c("subject_id","visit","batch")] <- do.call("rbind",sapply(as.character(sig.gsva.scores$sample),strsplit,"_"))
sig.gsva.scores$leadingEdge.size <- sapply(sig.gsva.scores$leadingEdge,function(x){length(unlist(strsplit(x," ")))}) 
sig.gsva.scores$pct.leadingEdge <- sig.gsva.scores$leadingEdge.size/sig.gsva.scores$size

pathway.info[sig.gsva.scores$cell.type.pathway,c("num.included","leadingEdge.size","pct.leadingEdge","leadingEdge")] <- 
  sig.gsva.scores[,c("size","leadingEdge.size","pct.leadingEdge","leadingEdge")]
```
  
### Pathway Selection
Select pathways correlated with DSM (using Spearman correlation)
```{r network,out.width="800px",fig.height=8}
sig.gsva.scores <- reshape2::dcast(sig.gsva.scores, subject_id + visit ~ pathway + celltype,value.var = "module.score",fun.aggregate = mean)
sig.gsva.scores <- sig.gsva.scores[order(sig.gsva.scores$visit),]
t0.sig.gsva.scores <- sig.gsva.scores[!duplicated(sig.gsva.scores$subject_id),]
#t0.sig.gsva.scores <- cbind(t0.sig.gsva.scores,end.point.info[t0.sig.gsva.scores$Donor,grep("corrected",colnames(end.point.info))])
rownames(t0.sig.gsva.scores) <- t0.sig.gsva.scores$subject_id
t0.sig.gsva.scores <- subset(t0.sig.gsva.scores,visit == "T0")
t0.sig.gsva.scores <- t0.sig.gsva.scores[,-c(1:2)]
# remove healthy?
#t0.sig.gsva.scores <- t0.sig.gsva.scores[grep("^HGR",rownames(t0.sig.gsva.scores)),]

# correlation between PC1 and module scores
t0.sig.gsva.scores$PC1 <- end.point.info[rownames(t0.sig.gsva.scores),"PC1"]
t0.sig.gsva.scores <- subset(t0.sig.gsva.scores,!is.na(PC1))
module.cor.res <- data.frame()
for (i in setdiff(colnames(t0.sig.gsva.scores),"PC1")) {
  tmp <- cor.test(t0.sig.gsva.scores[,"PC1"],t0.sig.gsva.scores[,i],method = "spearman")
  module.cor.res <- rbind(module.cor.res,data.frame(module=i,cor=tmp$estimate,cor.pval=tmp$p.value))
}

# correlation among modules with sig. correlation to PC1
sig.modules <- subset(module.cor.res,cor.pval <= 0.05)$module
cat("Number of modules sig. correlated with PC1:",length(sig.modules),"\n")
sig.modules <- c(sig.modules,"PC1")
sig.module.cor <- Hmisc::rcorr(as.matrix(t0.sig.gsva.scores[,sig.modules]),type = "spearman")
sig.module.cor.pval <- sig.module.cor$P
sig.module.cor <- sig.module.cor$r
pheatmap(sig.module.cor,show_rownames = T,show_colnames = F,main="Correlation of PC1-associated gene sets (excluding healthy)",
         color = viridis::viridis_pal()(101),fontsize = 8,
         annotation_row = pathway.info[,c("celltype","Category","PC1continuous_NES","PC1discrete_NES")])
```

### Inverse covariance matrix
Use the pathways selected in the previous section as input to calculate inverse covariance matrix.

Identify primary nodes, which are direct correlates of PC1, and secondary nodes, which are nodes connecting to primary nodes but not PC1
```{r icm}
# inverse covariance matrix method
inverse.matrix.method = "mb"
# use variance matrix as input
huge.transformed.matrix <- var(scale(as.matrix(t0.sig.gsva.scores[,sig.modules])),use = "pairwise.complete.obs")
#huge.transformed.matrix <- huge.npn(as.matrix(t0.sig.gsva.scores[,sig.modules])) 

lasso.graphs <- huge(huge.transformed.matrix,method=inverse.matrix.method,lambda = seq(1,0,-0.01))
num.connections <- sapply(lasso.graphs$path,sum)
names(num.connections) <- lasso.graphs$lambda
plot(lasso.graphs$lambda,num.connections,type="l",main = paste0(inverse.matrix.method," (without transformation)"),
     xlab="lambda",ylab="Number of non-zeros in precision matrix")
#cat("Number of remaining connections with different penalties:\n")
#print(num.connections)
# PC1 connections across lambda range
PC1.connections <- sapply(lasso.graphs$path,function(x){x[colnames(huge.transformed.matrix) == "PC1",]})
rownames(PC1.connections) <- colnames(huge.transformed.matrix)
colnames(PC1.connections) <- round(lasso.graphs$lambda,3)
pheatmap(PC1.connections,show_colnames = T,cluster_cols = F,color = c("white","darkred"),cellwidth = 4,
         main = paste0("Gene sets connected to PC1 (",inverse.matrix.method," without transformation)"),fontsize = 5)

inverse.matrix.penalty <- 0.35
sig.module.inverse.cov <- as.matrix(lasso.graphs$path[[which(abs(lasso.graphs$lambda - inverse.matrix.penalty) < 1e-6)]]) # some slight numerical instability
colnames(sig.module.inverse.cov) <- rownames(sig.module.inverse.cov) <- colnames(huge.transformed.matrix)
diag(sig.module.inverse.cov) <- 0
# nodes connected to PC1
primary.nodes <- names(which(sig.module.inverse.cov["PC1",] == 1))
# secondary nodes
secondary.nodes <- rownames(which(sig.module.cor.pval[,primary.nodes] <= 0.05 & sig.module.inverse.cov[,primary.nodes] == 1,arr.ind = T))
secondary.nodes <- setdiff(unique(secondary.nodes),c("PC1",primary.nodes))
```

### Create Network
Draw a simple network connecting PC1/DSM to the adjacent nodes 
```{r visualize.network,out.width="800px",fig.height=8}
network.nodes <- c("PC1",primary.nodes,secondary.nodes)
network.node.cor <- sig.module.cor[network.nodes,network.nodes]
network.node.cor.pval <- sig.module.cor.pval[rownames(network.node.cor),colnames(network.node.cor)]
network.node.inverse.cov <- sig.module.inverse.cov[rownames(network.node.cor),colnames(network.node.cor)]
network.node.cor.pval[is.na(network.node.cor.pval)] <- 1

pathway.info$num.connections <- NA
pathway.info[rownames(network.node.inverse.cov),]$num.connections <- rowSums(network.node.inverse.cov*(network.node.cor.pval <= 0.05),na.rm = T)
celltype.connection.counts <- aggregate(network.node.inverse.cov*(network.node.cor.pval <= 0.05),list(pathway.info[rownames(network.node.inverse.cov),"celltype"]),sum)
pathway.info$num.connections.celltype <- NA
pathway.info[colnames(celltype.connection.counts)[-1],]$num.connections.celltype <- colSums(celltype.connection.counts[,-1] > 0)

adj.matrix <- network.node.cor
# simple network for 1st level nodes only
#adj.matrix <- network.node.cor[unique(c("PC1",primary.nodes)),unique(c("PC1",primary.nodes))]
# remove connections that don't have sig. correlations
adj.matrix[network.node.cor.pval[rownames(adj.matrix),colnames(adj.matrix)] > 0.05] <- 0
cat("Number of connections remaining:",sum(adj.matrix != 0),"\n")
# remove connections not in inverse covariance matrix
adj.matrix[network.node.inverse.cov[rownames(adj.matrix),colnames(adj.matrix)] == 0] <- 0
cat("Number of connections remaining:",sum(adj.matrix != 0),"\n")
cat("Number of connections remaining:",sum(adj.matrix != 0),"\n")
# remove orphans
#adj.matrix <- adj.matrix[rowSums(abs(adj.matrix)) > 0,colSums(abs(adj.matrix)) > 0]

# edge type: 1 - PC1 to primary nodes, 2 - primary to secondary, 3 - same level connections
edge.type <- matrix(3,nrow=nrow(adj.matrix),ncol=ncol(adj.matrix),dimnames = dimnames(adj.matrix))
edge.type["PC1",] <- 1
edge.type[,"PC1"] <- 1
edge.type[primary.nodes,secondary.nodes] <- 2
edge.type[secondary.nodes,primary.nodes] <- 2

selected.rep.pathways <- pathway.info[rownames(adj.matrix),]
selected.rep.pathways$pathway <- rownames(adj.matrix)
selected.rep.pathways[is.na(selected.rep.pathways$pct.leadingEdge),"pct.leadingEdge"] <- max(selected.rep.pathways$pct.leadingEdge,na.rm = T)
rownames(selected.rep.pathways) <- selected.rep.pathways$pathway
# organize by node level then cell type
selected.rep.pathways$level <- 0
selected.rep.pathways[primary.nodes,"level"] <- 1
selected.rep.pathways[secondary.nodes,"level"] <- 2
selected.rep.pathways <- selected.rep.pathways[order(selected.rep.pathways$level,selected.rep.pathways$celltype),]
adj.matrix <- adj.matrix[rownames(selected.rep.pathways),rownames(selected.rep.pathways)]
net <- network(adj.matrix,directed = F)
network.vertex.names(net) <- rownames(adj.matrix)
cluster.color <- ggsci::pal_igv()(length(unique(selected.rep.pathways$category)))
names(cluster.color) <- unique(selected.rep.pathways$category) 
celltype.color <- c(ggsci::pal_d3()(10),rev(ggsci::pal_tron()(7)))
celltype.color <- celltype.color[1:length(unique(selected.rep.pathways$celltype))]
names(celltype.color) <- unique(selected.rep.pathways$celltype)  
# node attributes
net %v% "cluster" = as.character(selected.rep.pathways$Category)
net %v% "celltype" <- as.character(selected.rep.pathways$celltype)
net %v% "pctLeading" <- selected.rep.pathways$pct.leadingEdge
# edge attributes
set.edge.attribute(net,"direction",1)
set.edge.value(net,"direction",adj.matrix > 0)
set.edge.attribute(net,"correlation",1)
set.edge.value(net,"correlation",round(adj.matrix,3))
set.edge.attribute(net,"corr.pval",1)
set.edge.value(net,"corr.pval",network.node.cor.pval[rownames(adj.matrix),colnames(adj.matrix)])
set.edge.attribute(net,"type",1)
set.edge.value(net,"type",edge.type)

# overlap of leading edge
network.node.leading.edges <- gsva.scores[match(c(primary.nodes,secondary.nodes),gsva.scores$cell.type.pathway),c("leadingEdge","size","cell.type.pathway")]
network.node.leading.genes <- sapply(network.node.leading.edges$leadingEdge,strsplit," ")
names(network.node.leading.genes) <- network.node.leading.edges$cell.type.pathway

network.table <- ggnetwork(net)
network.table <- cbind(network.table,pathway.info[network.table$vertex.names,c("Primary_gene_sets","num.connections","num.connections.celltype")])
network.table[is.na(network.table$celltype),c("celltype","Primary_gene_sets")] <- "PC1"
network.table$selected.leading.genes <- sapply(network.node.leading.genes[network.table$vertex.names],function(x){paste0(x[1:5],collapse="\n")})
network.table <- merge(network.table,unique(network.table[,-c(7:12)]),by.x=c("xend","yend"),by.y=c("x","y"),all.x=T,suffixes=c("",".end"))

# subset to only PC1 and primary nodes
display.network <- subset(network.table,vertex.names %in% c("PC1",primary.nodes) & vertex.names.end %in% c("PC1",primary.nodes))
ggplot(display.network,aes(x=x,y=y,xend=xend,yend=yend)) + #,layout="circle") + 
  geom_edges(data=subset(display.network,type==3),aes(linetype=direction,size=-log10(corr.pval)/100,color=as.factor(type)),curvature = 0.2,alpha=.75) + 
  geom_edges(data=subset(display.network,type==1),aes(linetype=direction,size=-log10(corr.pval)/100,color=as.factor(type)),curvature = 0.1,alpha=.75) + 
  geom_nodes(pch=21,color="black",fill="white",size=15) + # background
  geom_nodes(aes(fill=celltype,size=pctLeading),pch=21,color="black") + #,size=10) + 
  geom_nodetext_repel(aes(label=Primary_gene_sets),size=2,force=2,fontface="bold",segment.size = 0,nudge_y = -0.07) + 
  geom_nodetext_repel(aes(label=selected.leading.genes),size=2,force=2,nudge_y = 0.1,nudge_x = 0.1) +
  geom_edgetext(aes(label=correlation,color=direction),size=4) + 
  theme_blank() + scale_fill_manual(values=c("orange","lightblue","darkgreen")) + 
  scale_color_manual(values=c("1"="grey30","2"="deepskyblue3","3"="grey90","TRUE"="tomato","FALSE"="deepskyblue2")) + 
  scale_alpha(range=c(0.3,1)) + scale_size(range = c(5,10)) +#scale_size(range = c(2,6)) +
  scale_linetype_manual(values=c("dashed","solid")) + guides(color=F,linetype=F) + theme(plot.margin = unit(c(1,1,1,1),"inch"))

```

### Session Info
```{r save}
# save network
tmp <- subset(network.table,!is.na(correlation))
tmp <- tmp[order(tmp$type,tmp$celltype.end,tmp$vertex.names.end,tmp$celltype),]
write.csv(tmp,file.path(output.folder,"network.table.csv"),row.names=F)

sessionInfo()
```