---
title: "3.UMAPandProteinVis.Figure2"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide")
```

Load seurat object from GEO and make heatmap of average protein expression for Figure 2A:

```{r Figure2A, fig.height = 15, fig.width = 6}
#this is tested using R 3.6.1 on a high-performance comupting node with 8 cores and at least 320 gb or ram. 
library("Seurat") #load Seurat 3.1
library("dplyr")
library("matrixStats")
library('tidyverse')
library('parallelDist')
library("pheatmap")
library("viridis")
library("ggridges")
library("magrittr")
library("genefilter")
library("DescTools")
library("scico")
library("grid")

## import the seurat object downloaded from GEO
merge.SNG = readRDS("SeuratObjects/AllBatches_SeuratObj.rds")

##remove the CHI014 technical control sample
merge.SNG <- subset(merge.SNG, Donor == "CHI014", invert = TRUE)

## make a object of the cluster averages for heatmaps 
Idents(merge.SNG) <- "WCTcoursecelltype"
source("utilityFunctions/AverageExpression_MeanOnly.r")
environment(AverageExpression_MeanOnly) <- asNamespace('Seurat') # need this to allow the function to call hidden seurat package functions
merge.SNG.Aver = AverageExpression_MeanOnly(merge.SNG, assays = "limmaCITE")
quantile_breaks <- function(xs, n = 100) {
  breaks <- quantile(xs, probs = seq(0.5, 1, length.out = n))
  breaks[!duplicated(breaks)]
}

f2 <- pOverA(0.01, 3)
ffun2 <- filterfun(f2)
merge.SNGADTfilt = genefilter(GetAssayData(merge.SNG[["limmaCITE"]], slot = "data"), ffun2)

mat_breaks <- quantile_breaks(as.matrix(merge.SNG.Aver$limmaCITE[which(merge.SNGADTfilt),]), n = 101)
pheatmap(merge.SNG.Aver$limmaCITE[which(merge.SNGADTfilt),which(!(colnames(merge.SNG.Aver$limmaCITE) %in% c("Dblt","Unknown","gated_out","dim","Tcell","dblt")))], scale = "none", border_color=NA, viridis(n=100), breaks = mat_breaks)
```

Innate population UMAP for Figure 2B
```{r Figure2Binnate, fig.width = 6, fig.height = 6}
innate.merge.SNG <- subset(merge.SNG, subset = WCTcoursecelltype %in% 
                             c("Mono_Classical", "Mono_NonClassical", "NK_CD16hi",
                               "NK_CD56loCD16lo", "pDC", "cDC", "Platelets", "NK_CD56hiCD16lo",
                               "Mono_Intermediate", "Granulocytes"))
innate.merge.SNGADTfilt = genefilter(GetAssayData(innate.merge.SNG[["limmaCITE"]], slot = "data"), ffun2)
innate.merge.SNG <- RunUMAP(innate.merge.SNG, assay = "limmaCITE", features = rownames(innate.merge.SNG[["limmaCITE"]][which(innate.merge.SNGADTfilt),]),
                     n.neighbors = 15, min.dist = 0.01, spread =  5)
innate.p <- DimPlot(innate.merge.SNG, 
                      group.by = "WCTcoursecelltype", label = TRUE) + NoLegend()
innate.p <- AugmentPlot(innate.p, width = 6, height = 6)
innate.p

```

Adaptive population UMAP for Figure 2B
```{r Figure2Badaptive, fig.width = 6, fig.height = 6}

adaptive.merge.SNG <- subset(merge.SNG, subset = WCTcoursecelltype %in% 
                             c("B_Naive", "B_Mem", "gammadeltaT",
                               "PB_Plasmablasts", "Treg", "CD4_Mem", "CD8_Mem", "CD8_Naive",
                               "CD4_Naive", "MAIT", "TissueResMemT", "DPT", "DNT", "TCRVbeta13.1pos"))
adaptive.merge.SNGADTfilt = genefilter(GetAssayData(adaptive.merge.SNG[["limmaCITE"]], slot = "data"), ffun2)
adaptive.merge.SNG <- RunUMAP(adaptive.merge.SNG, assay = "limmaCITE", features = rownames(adaptive.merge.SNG[["limmaCITE"]][which(adaptive.merge.SNGADTfilt),]),
                            n.neighbors = 15, min.dist = 0.01, spread =  5)
adaptive.p <- DimPlot(adaptive.merge.SNG, 
                      group.by = "WCTcoursecelltype", label = TRUE) + NoLegend()
adaptive.p <- AugmentPlot(adaptive.p, width = 6, height = 6)
adaptive.p
```

Average protein expression heatmap of selected T cell populations for Figure 2C
```{r Figure2C, fig.width = 4, fig.height = 8}
Idents(merge.SNG) <- "WCTmergedcelltype"
CD4_Mem.Aver = AverageExpression_MeanOnly(subset(merge.SNG, WCTcoursecelltype == "CD4_Mem"), assay = "limmaCITE")
f1 <- pOverA(0.01, 2)
ffun <- filterfun(f1)
CD4_Mem.ADTfilt = genefilter(CD4_Mem.Aver$limmaCITE[,c("CD4_Mem_TM","CD4_Mem_EM.TE","CD4_Mem_CM")], ffun)
#mat_breaks <- quantile_breaks(as.matrix(CD4_Mem.Aver$limmaCITE[which(CD4_Mem.ADTfilt),]), n = 101)
mat_breaks <- quantile_breaks(as.matrix(merge.SNG.Aver$limmaCITE[which(merge.SNGADTfilt),]), n = 101)

pheatmap(CD4_Mem.Aver$limmaCITE[which(CD4_Mem.ADTfilt),c("CD4_Mem_TM","CD4_Mem_EM.TE","CD4_Mem_CM")], scale = "none", border_color=NA, viridis(n=100), breaks = mat_breaks)

```



Mono_Classical population UMAP for Figure 2D, colored by COVID or healthy control
```{r Figure2D, fig.width = 6, fig.height = 6}
Mono_Classical <- subset(merge.SNG, subset = WCTcoursecelltype %in% c( "Mono_Classical"))
Mono_Classical <- FindVariableFeatures(Mono_Classical, assay = "RNA")
Mono_Classical <- ScaleData(Mono_Classical)
Mono_Classical <- RunPCA(Mono_Classical)
Mono_Classical <- RunUMAP(Mono_Classical, dims = 1:15)
MonoUMAP.Class <- DimPlot(Mono_Classical, group.by = "Class", cols = c("#00BA38", "#F8766D"))
MonoUMAP.Class <- AugmentPlot(MonoUMAP.Class, width = 6, height = 6)
MonoUMAP.Class
```
Mono_Classical population UMAP for Figure 2E, colored by DSM (labelled PC1 here) class
```{r Figure2E, fig.width = 6, fig.height = 6}
MonoUMAP.DSM_cat <- DimPlot(subset(Mono_Classical, PC1_cat %in% c("PC1_low","PC1_high")), group.by = "PC1_cat",cols = c("#619CFF", "#F8766D"))
MonoUMAP.DSM_cat <- AugmentPlot(MonoUMAP.DSM_cat, width = 6, height = 6)
MonoUMAP.DSM_cat
```
Mono_Classical population UMAP for Figure 2F, colored by days since symptom onset
```{r Figure2F, fig.width = 6, fig.height = 6}
MonoUMAP.dayssinceonset <- FeaturePlot(subset(Mono_Classical, PC1_cat %in% c("PC1_low","PC1_high") & days_since_onset != "NA"), features = "days_since_onset")
MonoUMAP.dayssinceonset <- AugmentPlot(MonoUMAP.dayssinceonset, width = 6, height = 6)
MonoUMAP.dayssinceonset
```
Mono_Classical population UMAP for Figure 2G, CD163-high monocytes in black
```{r Figure2G, fig.width = 6, fig.height = 6}
MonoUMAP.CD163 <- DimPlot(subset(Mono_Classical, WCTmergedcelltype %in% c("Mono_Classical","Mono_Classical_CD163hi")), group.by = "WCTmergedcelltype", cols = c("grey","black"))
MonoUMAP.CD163 <- AugmentPlot(MonoUMAP.CD163, width = 6, height = 6)
MonoUMAP.CD163
```
Mono_Classical population UMAP for Figure 2G (alternate), CD163 protein expression
```{r Figure2Galt, fig.width = 6, fig.height = 6}
MonoUMAP.CD163x <- FeaturePlot(subset(Mono_Classical, WCTmergedcelltype %in% c("Mono_Classical","Mono_Classical_CD163hi")), features = "cite_CD163",max.cutoff = 7.5 )
MonoUMAP.CD163x <- AugmentPlot(MonoUMAP.CD163x, width = 6, height = 6)
MonoUMAP.CD163x
```
Mono_Classical population UMAP for Figure 2H, HLA-DR protein expression
```{r Figure2H, fig.width = 6, fig.height = 6}
MonoUMAP.HLADRx <- FeaturePlot(subset(Mono_Classical, WCTmergedcelltype %in% c("Mono_Classical","Mono_Classical_CD163hi")), features = "cite_HLA-DR",max.cutoff = 7.5)
MonoUMAP.HLADRx <- AugmentPlot(MonoUMAP.HLADRx, width = 6, height = 6)
MonoUMAP.HLADRx
```
Mono_Classical population UMAP for Figure 2I, colored by Donor
```{r Figure2I, fig.width = 6, fig.height = 6}
MonoUMAP.Donor <- DimPlot(Mono_Classical, group.by = "Donor") + NoLegend()
MonoUMAP.Donor <- AugmentPlot(MonoUMAP.Donor, width = 6, height = 6)
MonoUMAP.Donor
```


```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```

