---
title: "4.TcellReclust.CD8"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide")
```

## Recluster on RNA within the T cell subsets

Load R packages and data:
```{r loadpackagesanddata}
#this is tested using R 3.6.1 on a high-performance comupting node with 8 cores and at least 320 gb or ram. 
library("Seurat") #load Seurat 3.1
library("dplyr")
library("matrixStats")
library('tidyverse')
library('parallelDist')
library("pheatmap")
library("viridis")
library("ggridges")
library("magrittr")
library("genefilter")
library("DescTools")
library("scico")

#### Recluster on RNA within the T cell subsets ####

## import the seurat object downloaded from GEO
merge.SNG = readRDS("SeuratObjects/AllBatches_SeuratObj.rds")

## import fgsea results tables
t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable = read.table("Metadata/2020_08_09_t0_plus_healthy--healthy_vs_covid--COVID-Healthy.tsv", header = TRUE, sep = "\t")
t0_plus_t0_covid_only.PC1.fgseaatable = read.table("Metadata/2020_08_09_t0_covid_only--PC1--PC1.tsv", header = TRUE, sep = "\t")

```

# Recluster within the CD8 cells, using the geneset enrichment leading edge genes identified in the pseudobulk differential expression analysis

```{r reclustering, echo=FALSE}
### RNA clustering
library(genefilter)
f1 <- pOverA(0.01, 2)
ffun <- filterfun(f1)

dir.create("withinWCTCelltypeRNAclustPathLeadingEdge")
Pathway.leadingEdge = union(unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable$leadingEdge),  " "))),
                              unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$leadingEdge),  " ")))
                            )
#####  cluster RNA within the Total CD8

#filter out the cell subsets of interest
TotalCD8.LG = subset(merge.SNG, WCTcoursecelltype %in% c("CD8_Naive","CD8_Mem"))

#select just the leading edge genes for the pathway enrichment results of the relevant cell subsets
tO.COVIDvsHealthy.SelectedPath.leadingEdge.LG.CD8 = unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable[
  which(
    t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$pathway %in% c(
      "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
      "HALLMARK_INFLAMMATORY_RESPONSE",
      "btm_M4.1_cell cycle (I)",
      "GO_RESPONSE_TO_TYPE_I_INTERFERON",
      "KEGG_RIBOSOME",
      "KEGG_GLYCOLYSIS_GLUCONEOGENESIS",
      "KEGG_OXIDATIVE_PHOSPHORYLATION",
      "reactome_Fatty acid metabolism"
    ) 
    &
      t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$celltype %in% c(
        "CD8_Naive",
        "CD8_Mem"
      )  
  ),]$leadingEdge),  " ")))
tO.PC1.SelectedPath.leadingEdge.LG.CD8 = unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable[
  which(
    t0_plus_t0_covid_only.PC1.fgseaatable$pathway %in% c(
      "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
      "HALLMARK_INFLAMMATORY_RESPONSE",
      "btm_M4.1_cell cycle (I)",
      "GO_RESPONSE_TO_TYPE_I_INTERFERON",
      "KEGG_RIBOSOME",
      "KEGG_GLYCOLYSIS_GLUCONEOGENESIS",
      "KEGG_OXIDATIVE_PHOSPHORYLATION",
      "reactome_Fatty acid metabolism"
    ) 
    &
      t0_plus_t0_covid_only.PC1.fgseaatable$celltype %in% c(
        "CD8_Naive",
        "CD8_Mem"
      )  
  ),]$leadingEdge),  " ")))

#run the clustering within the selected subset using filtered genes
TotalCD8.LG = ScaleData(TotalCD8.LG, assay="RNA", features = unique(c(tO.COVIDvsHealthy.SelectedPath.leadingEdge.LG.CD8,tO.PC1.SelectedPath.leadingEdge.LG.CD8)), vars.to.regress=c("Donor","Batch"))
TotalCD8.LG <- RunPCA(TotalCD8.LG, assay="RNA", slot = "scale.data",  features = unique(c(tO.COVIDvsHealthy.SelectedPath.leadingEdge.LG.CD8,tO.PC1.SelectedPath.leadingEdge.LG.CD8)), reduction.name = "pcaledge")
TotalCD8.LG <- FindNeighbors(TotalCD8.LG, dims = 1:15, reduction = "pcaledge")
TotalCD8.LG <- FindClusters(TotalCD8.LG, resolution = c(1), algorithm = 1)


```

#Plot clusters for figure 5
Figure 5A:
```{r Figure5A,  fig.width = 10, fig.height = 10}
library(ComplexHeatmap)
library(viridis)
viridis(9)

### heatmap for the CD8 clusters on leading edge genes of selected pathways
TotalCD8.LG$WCTRNAclusters = paste0("TotalCD8.LG_",TotalCD8.LG$RNA_snn_res.1)
TotalCD8.LG = ScaleData(TotalCD8.LG, assay="RNA", features = unique(c(tO.COVIDvsHealthy.SelectedPath.leadingEdge.LG.CD8,tO.PC1.SelectedPath.leadingEdge.LG.CD8)))
TotalCD8.selectclust.LG = subset(TotalCD8.LG, Sorted == "N" & RNA_snn_res.1 %in% c(14) & Timepoint == "T0")
# also label genesets
inflamgenes <- rownames(GetAssayData(TotalCD8.selectclust.LG, assay = "RNA", slot = "scale.data")) %in% 
  union(unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable[
    which(
      t0_plus_t0_covid_only.PC1.fgseaatable$pathway %in% c(
        
        "HALLMARK_INFLAMMATORY_RESPONSE"
        
      ) 
      &
        t0_plus_t0_covid_only.PC1.fgseaatable$celltype %in% c(
          "CD8_Naive",
          "CD8_Mem"
        )  
    ),]$leadingEdge),  " ")))
    ,
    unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable[
      which(
        t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$pathway %in% c(
          
          "HALLMARK_INFLAMMATORY_RESPONSE"
          
        ) 
        &
          t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$celltype %in% c(
            "CD8_Naive",
            "CD8_Mem"
          )  
      ),]$leadingEdge),  " ")))
  )

cellcyclegenes <- rownames(GetAssayData(TotalCD8.selectclust.LG, assay = "RNA", slot = "scale.data")) %in% 
  union(unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable[
    which(
      t0_plus_t0_covid_only.PC1.fgseaatable$pathway %in% c(
        
        "btm_M4.1_cell cycle (I)"
      ) 
      &
        t0_plus_t0_covid_only.PC1.fgseaatable$celltype %in% c(
          "CD8_Naive",
          "CD8_Mem"
        )  
    ),]$leadingEdge),  " ")))
    ,
    unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable[
      which(
        t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$pathway %in% c(
          
          "btm_M4.1_cell cycle (I)"
          
        ) 
        &
          t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$celltype %in% c(
            "CD8_Naive",
            "CD8_Mem"
          )  
      ),]$leadingEdge),  " ")))
  )

T1INFgenes <- rownames(GetAssayData(TotalCD8.selectclust.LG, assay = "RNA", slot = "scale.data")) %in% 
  union(unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable[
    which(
      t0_plus_t0_covid_only.PC1.fgseaatable$pathway %in% c(
        
        "GO_RESPONSE_TO_TYPE_I_INTERFERON"
      ) 
      &
        t0_plus_t0_covid_only.PC1.fgseaatable$celltype %in% c(
          "CD8_Naive",
          "CD8_Mem"
        )  
    ),]$leadingEdge),  " ")))
    ,
    unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable[
      which(
        t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$pathway %in% c(
          
          "GO_RESPONSE_TO_TYPE_I_INTERFERON"
        ) 
        &
          t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$celltype %in% c(
            "CD8_Naive",
            "CD8_Mem"
          )  
      ),]$leadingEdge),  " ")))
  )

HALLMARK_TNFA_SIGNALING_VIA_NFKB <- rownames(GetAssayData(TotalCD8.selectclust.LG, assay = "RNA", slot = "scale.data")) %in% 
  union(unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable[
    which(
      t0_plus_t0_covid_only.PC1.fgseaatable$pathway %in% c(
        "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
        
      ) 
      &
        t0_plus_t0_covid_only.PC1.fgseaatable$celltype %in% c(
          "CD8_Naive",
          "CD8_Mem"
        )  
    ),]$leadingEdge),  " ")))
    ,
    unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable[
      which(
        t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$pathway %in% c(
          "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
        ) 
        &
          t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$celltype %in% c(
            "CD8_Naive",
            "CD8_Mem"
          )  
      ),]$leadingEdge),  " ")))
  )

ribosome <- rownames(GetAssayData(TotalCD8.selectclust.LG, assay = "RNA", slot = "scale.data")) %in% 
  union(unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable[
    which(
      t0_plus_t0_covid_only.PC1.fgseaatable$pathway %in% c(
        
        "KEGG_RIBOSOME"
      ) 
      &
        t0_plus_t0_covid_only.PC1.fgseaatable$celltype %in% c(
          "CD8_Naive",
          "CD8_Mem"
        )  
    ),]$leadingEdge),  " ")))
    ,
    unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable[
      which(
        t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$pathway %in% c(
          
          "KEGG_RIBOSOME"
        ) 
        &
          t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$celltype %in% c(
            "CD8_Naive",
            "CD8_Mem"
          )  
      ),]$leadingEdge),  " ")))
  )

KEGG_GLYCOLYSIS_GLUCONEOGENESIS <- rownames(GetAssayData(TotalCD8.selectclust.LG, assay = "RNA", slot = "scale.data")) %in% 
  union(unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable[
    which(
      t0_plus_t0_covid_only.PC1.fgseaatable$pathway %in% c(
        
        "KEGG_GLYCOLYSIS_GLUCONEOGENESIS"
      ) 
      &
        t0_plus_t0_covid_only.PC1.fgseaatable$celltype %in% c(
          "CD8_Naive",
          "CD8_Mem"
        )  
    ),]$leadingEdge),  " ")))
    ,
    unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable[
      which(
        t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$pathway %in% c(
          
          "KEGG_GLYCOLYSIS_GLUCONEOGENESIS"
        ) 
        &
          t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$celltype %in% c(
            "CD8_Naive",
            "CD8_Mem"
          )  
      ),]$leadingEdge),  " ")))
  )

KEGG_OXIDATIVE_PHOSPHORYLATION <- rownames(GetAssayData(TotalCD8.selectclust.LG, assay = "RNA", slot = "scale.data")) %in% 
  union(unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable[
    which(
      t0_plus_t0_covid_only.PC1.fgseaatable$pathway %in% c(
        
        "KEGG_OXIDATIVE_PHOSPHORYLATION"
      ) 
      &
        t0_plus_t0_covid_only.PC1.fgseaatable$celltype %in% c(
          "CD8_Naive",
          "CD8_Mem"
        )  
    ),]$leadingEdge),  " ")))
    ,
    unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable[
      which(
        t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$pathway %in% c(
          
          "KEGG_OXIDATIVE_PHOSPHORYLATION"
        ) 
        &
          t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$celltype %in% c(
            "CD8_Naive",
            "CD8_Mem"
          )  
      ),]$leadingEdge),  " ")))
  )

reactome_Fattyacidmetabolism <- rownames(GetAssayData(TotalCD8.selectclust.LG, assay = "RNA", slot = "scale.data")) %in% 
  union(unique(unlist(strsplit(as.character(t0_plus_t0_covid_only.PC1.fgseaatable[
    which(
      t0_plus_t0_covid_only.PC1.fgseaatable$pathway %in% c(
        
        "reactome_Fatty acid metabolism"
      ) 
      &
        t0_plus_t0_covid_only.PC1.fgseaatable$celltype %in% c(
          "CD8_Naive",
          "CD8_Mem"
        )  
    ),]$leadingEdge),  " ")))
    ,
    unique(unlist(strsplit(as.character(t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable[
      which(
        t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$pathway %in% c(
          
          "reactome_Fatty acid metabolism"
        ) 
        &
          t0_plus_healthy.healthy_vs_covid.COVID.Healthy.fgseaatable$celltype %in% c(
            "CD8_Naive",
            "CD8_Mem"
          )  
      ),]$leadingEdge),  " ")))
  )

ha.CD8 = HeatmapAnnotation(
  #PC1 = TotalCD8.selectclust.LG$PC1,
  #PC1class = TotalCD8.selectclust.LG$PC1_cat,
  #Class = TotalCD8.selectclust.LG$Class,
  Cluster = TotalCD8.selectclust.LG$WCTRNAclusters
)


library("circlize")
col_fun = colorRamp2(c(-1.5, 0, 1.5), c("#FF00FF", "#000000", "#FFFF00"))


Idents(TotalCD8.LG) <- "RNA_snn_res.1"
TotalCD8.T0.LG = subset(TotalCD8.LG, Sorted == "N" & Timepoint == "T0")
TotalCD8.T0.LG = subset(TotalCD8.T0.LG, idents = c("16","17"), invert = TRUE)
TotalCD8.T0.aver.LG = AverageExpression(TotalCD8.T0.LG, return.seurat = TRUE)
TotalCD8.T0.aver.LG = ScaleData(TotalCD8.T0.aver.LG, assay="RNA", features = unique(c(tO.COVIDvsHealthy.SelectedPath.leadingEdge.LG.CD8,tO.PC1.SelectedPath.leadingEdge.LG.CD8)))

GOI = c("MX1","IFNAR1","IFNGR2","IRF3", "JUNB","PCNA","CEBPB","SOCS3","IL23A","EIF1","HIF1A","ALDOA", "POLA1", "HELLS", "PRC1",
        "STAT2","OAS1","NFKBIA","IRF7","MKI67", "TNFAIP3","TNF","TAP1","CD69","RELA", "SELL" ,"ISG15","AREG","RPL13A","RPL22")
label <- which(rownames(GetAssayData(TotalCD8.T0.aver.LG, assay = "RNA", slot = "scale.data")) %in% GOI)
label_genes <- rownames(GetAssayData(TotalCD8.T0.aver.LG, assay = "RNA", slot = "scale.data"))[label]

Heatmap(as.matrix(GetAssayData(TotalCD8.T0.aver.LG, assay = "RNA", slot = "scale.data")), name = "CD8", show_column_names = TRUE,
        cluster_columns = TRUE, column_title_rot = 90,
        #column_split = TotalCD8.T0.aver.LG$,
        col = col_fun,
        row_km = 8,
        row_names_gp = gpar(fontsize = 10)
) +
  Heatmap(inflamgenes + 0, name = "HALLMARK_INFLAMMATORY_RESPONSE", col = c("0" = "white", "1" = "purple"),
          show_heatmap_legend = FALSE, width = unit(2, "mm"), column_names_gp = gpar(fontsize = 6)) +
  Heatmap(HALLMARK_TNFA_SIGNALING_VIA_NFKB + 0, name = "HALLMARK_TNFA_SIGNALING_VIA_NFKB", col = c("0" = "white", "1" = "red"),
          show_heatmap_legend = FALSE, width = unit(2, "mm"), column_names_gp = gpar(fontsize = 6)) +
  Heatmap(cellcyclegenes + 0, name = "btm_M4.1_cellcycle", col = c("0" = "white", "1" = "blue"),
          show_heatmap_legend = FALSE, width = unit(2, "mm"), column_names_gp = gpar(fontsize = 6)) +
  Heatmap(T1INFgenes + 0, name = "GO_RESPONSE_TO_TYPE_I_INTERFERON", col = c("0" = "white", "1" = "black"),
          show_heatmap_legend = FALSE, width = unit(2, "mm"), column_names_gp = gpar(fontsize = 6)) +
  Heatmap(ribosome + 0, name = "KEGG_RIBOSOME", col = c("0" = "white", "1" = "orange"),
          show_heatmap_legend = FALSE, width = unit(2, "mm"), column_names_gp = gpar(fontsize = 6)) +
  Heatmap(KEGG_GLYCOLYSIS_GLUCONEOGENESIS + 0, name = "KEGG_GLYCOLYSIS_GLUCONEOGENESIS", col = c("0" = "white", "1" = "pink"),
          show_heatmap_legend = FALSE, width = unit(2, "mm"), column_names_gp = gpar(fontsize = 6)) +
  Heatmap(KEGG_OXIDATIVE_PHOSPHORYLATION + 0, name = "KEGG_OXIDATIVE_PHOSPHORYLATION", col = c("0" = "white", "1" = "green"),
          show_heatmap_legend = FALSE, width = unit(2, "mm"), column_names_gp = gpar(fontsize = 6)) +
  Heatmap(reactome_Fattyacidmetabolism + 0, name = "reactome_Fatty acid metabolism", col = c("0" = "white", "1" = "grey"),
          show_heatmap_legend = FALSE, width = unit(2, "mm"), column_names_gp = gpar(fontsize = 6)) +
  rowAnnotation(foo = anno_mark(at = label,
                              labels = label_genes,
                              labels_gp = gpar(fontsize = 8)))


```

Barplot of fraction clonal cells per cluster for figure 5A:
```{r Figure5Apart2,  fig.width = 8, fig.height = 4}
expandedT = read.table("Metadata/2020_07_28_tcell_expansion_df.tsv", header = TRUE)

TotalCD8.LG$Expanded <- ifelse(colnames(TotalCD8.LG) %in% 
                                    expandedT$barcodeBatch[which(expandedT$expanded == TRUE)], TRUE, FALSE)

TotalCD8.T0.LG$Expanded <- ifelse(colnames(TotalCD8.T0.LG) %in% 
                                    expandedT$barcodeBatch[which(expandedT$expanded == TRUE)], TRUE, FALSE)

TotalCD8.T0.LG.unsort <- subset(TotalCD8.T0.LG, Sorted == "N")

#make barplots of expanded cell percentage per cluster
barplot(prop.table(table(TotalCD8.T0.LG.unsort$RNA_snn_res.1, TotalCD8.T0.LG.unsort$Expanded), 
                   margin = 1)[c(15,12,14,11,1,5,4,2,8,13,7,3,10,16,6,9),2],
        horiz = FALSE,
        ylim = c(0,1)
        )

```

Protein heatmap for figure 5B
```{r Figure5B,  fig.width = 8, fig.height = 4}
quantile_breaks <- function(xs, n = 100) {
  breaks <- quantile(xs, probs = seq(0.5, 1, length.out = n))
  breaks[!duplicated(breaks)]
}
source("utilityFunctions/AverageExpression_MeanOnly.r")
environment(AverageExpression_MeanOnly) <- asNamespace('Seurat') # need this to allow the function to call hidden seurat package functions
TotalCD8.T0.LGAver = AverageExpression_MeanOnly(TotalCD8.T0.LG, return.seurat=F)
mat_breaks.temp <- quantile_breaks(as.matrix(TotalCD8.T0.LGAver$limmaCITE[c("CD38", "HLA-DR","CD278","CD279", "CD86","KLRG1", "CD28","CD45RA","CD45RO"),]), n = 101)
pheatmap(TotalCD8.T0.LGAver$limmaCITE[c("CD38", "HLA-DR","CD278","CD279", "CD86","KLRG1", "CD28","CD45RA","CD45RO"),], scale = "none", border_color=NA, viridis(n=100), breaks = mat_breaks.temp)
```

Heatmap of overlap of RNA clusters and protein based cluster labels, for Figure 5D
```{r Fig5D, fig.width = 4, fig.height = 6,}
CD8.T0.LG.WCTmergedPerCluster = prop.table(table(Idents(TotalCD8.T0.LG), TotalCD8.T0.LG$WCTmergedcelltype), margin = 1)
pheatmap(CD8.T0.LG.WCTmergedPerCluster[,which(!colnames(CD8.T0.LG.WCTmergedPerCluster) %in% 
                                             c("CD8_Mem_CD158hi","CD8_Naive_CD41hi",
                                               "CD8_Mem_CD41hi","CD8_Mem_CD69pos",
                                               "CD8_Naive_MAIT","CD8_Mem_IgG.IsoBinding",
                                               "CD4_Naive","CD8_Naive_NKT","CD8_Naive_CD101hi",
                                               "CD8_Mem_NKT","CD8_Mem_CD101hi"))],
         border_color = NA)

```
