---
title: "Geneset enrichment scores and cytokine correlation scatter"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE)
library(matrixStats)
library(plyr)
library(tidyverse)
library(ggplot2)
library(Biobase)
library(reshape2)
library(ggpubr)
library(gridExtra)
source("util_fun/gsvascore.dfwmeta.R")
```

## Geneset enrichment gsva scores and cytokine correlations
**Generate Figure 4 B,C,G,I,L**
```{r input, include=FALSE}
### get cytokine data
load("input/covid19.metadata.paper1.RData")
PC1 <- readRDS("input/citeseq.patient.end.points.RDS")
PC1$PC1class <-  ifelse(PC1$PC1 > median(PC1$PC1), "PC1_high", "PC1_low")
PC1$sample_id <- paste(PC1$subject_id, PC1$visit, sep = "_")

IL15 <-  filter(covid19.lab.results, grepl("IL-15",test_name)) %>%
  mutate(test_value_log10 = log10(test_value+0.01)) %>%
  mutate(sample_id = paste(subject_id, test_time_point.PBMC, sep = "_"))
IL15_T0 <- IL15 %>% 
  dplyr::filter(test_time_point.PBMC == "T0") %>% 
  dplyr::group_by(sample_id) %>% 
  dplyr::summarise_at(vars(test_value_log10), mean) %>%
  dplyr::left_join(PC1, by = "sample_id")


### get gsva list data
# input is the files from output of sample_gsva_dftoelist.R
PC1_gsva_esetlist <- readRDS("output/module_score_gsva/PC1_PC1_module_score_gsva_filtered_samples_genes_esetlist.rds")

combined_genesets <- c("HALLMARK_TNFA_SIGNALING_VIA_NFKB", 
                       "HALLMARK_INFLAMMATORY_RESPONSE",
                       "reactome_Fatty acid metabolism",
                       "KEGG_CHEMOKINE_SIGNALING_PATHWAY",
                       "GO_CELLULAR_RESPONSE_TO_INTERLEUKIN_1",
                       "HALLMARK_MTORC1_SIGNALING")

PC1score_list_df <- getgsvascore_list_df(PC1_gsva_esetlist, combined_genesets)
PC1t0score_list_df <- PC1score_list_df %>% filter(Timepoint == "T0",pc1_group != "HC") %>%
  mutate(sample_id = paste(Subject, Timepoint, sep = "_"))
PC1t0score_list_df <- PC1t0score_list_df %>% 
  left_join(IL15_T0[,c("sample_id","test_value_log10")], by = "sample_id") %>%
  mutate(severity_outcome = as.character(severity_outcome))
```


### NK_CD16hi fatty acid metabolism correlations with inflammatory pathways and circulating IL15 level
```{r, fig.width=9, fig.height=5}
PC1t0score_list_df_NK <- filter(PC1t0score_list_df, celltype == "NK_CD16hi") %>%
  melt(id = colnames(.)[-c(2,3,5,6,7,19)])

p <- ggplot(PC1t0score_list_df_NK, aes(x = value, y = `reactome_Fatty acid metabolism`)) +
  geom_point(shape=21,aes(fill=severity_outcome),size=3, color="white")+
  scale_fill_manual(name="Severity_outcome",values = c("#374E55FF","#DF8F44FF","#00A1D5FF","#B24745FF")) +
  geom_smooth(se = F,method = "lm") + 
  stat_cor(method = "pearson") +
  facet_wrap(~variable, scales = "free") +
  theme_bw()

p

# ggsave(filename = "FAvs.score.IL15.NK.pdf", plot = p, device = "pdf", width = 9, height = 5)
```


### NK_CD16hi fatty acid metabolism correlations with IFG mRNA level
input is the pseudobulk eset_list object from pseudobulk.normalize.esetlist.R script  
```{r, fig.width=5, fig.height=3.3}
# input is the pseudobulk eset_list object from pseudobulk.normalize.esetlist.R script
eset_list <- readRDS("output/dge_lists/pbulk_eset_list_normalized_WCTcourse_metafiltered.rds")

# take T0 only and remove technical CHI014 control
eset_list_T0 <- lapply(eset_list, function(dge){
  filter <- colnames(dge)[dge@phenoData@data$Timepoint %in% c("T0", "HC") & 
                            !is.na(dge@phenoData@data$PC1_cat) & 
                            dge@phenoData@data$Donor != "CHI014"]
  dge <- dge[,filter]
})

combined_genes <- c("IFNG")
exprs_list_df <- getgsvascore_list_df(eset_list_T0, combined_genesets = combined_genes)

exprs_list_df <- exprs_list_df %>% filter(Timepoint == "T0",pc1_group != "HC") %>%
  mutate(sample_id = paste(Subject, Timepoint, sep = "_"))

exprs_list_df_NK <- exprs_list_df %>%
  filter(celltype == "NK_CD16hi") %>%
  left_join(PC1t0score_list_df, by = c("sample_id","celltype"))

p <- ggplot(exprs_list_df_NK, aes(x = IFNG, y = `reactome_Fatty acid metabolism`)) +
  geom_point(shape=21,aes(fill=severity_outcome.x),size=3, color="white")+
  scale_fill_manual(name="Severity_outcome",values = c("#374E55FF","#DF8F44FF","#00A1D5FF","#B24745FF")) +
  geom_smooth(se = F,method = "lm") + 
  stat_cor(method = "pearson") +
  theme_bw()

p
# ggsave(filename = "FAvs.score.IL15.NK.pdf", plot = p, device = "pdf", width = 5, height = 3.3)

```



### circulating IL15 correlation with severity at T0
```{r, fig.width=5, fig.height=3.3}
#extract only CITEseq T0 samples
IL15_T0 <- filter(IL15_T0, !is.na(subject_id)) %>%
  filter(sample_id %in% unique(PC1t0score_list_df$sample_id))
IL15_T0$severity.outcome <- paste(IL15_T0$severity, IL15_T0$outcome, sep="-")
p <- ggplot(IL15_T0, aes(x = test_value_log10, y = PC1)) +
  geom_point(shape=21,aes(fill=severity.outcome),size=3, color="white")+
  scale_fill_manual(name="Severity_outcome",values = c("#374E55FF","#DF8F44FF","#00A1D5FF","#B24745FF"))+
  geom_smooth(se = F,method = "lm") + 
  stat_cor(method = "pearson") +
  theme_bw()

p
# ggsave(filename = "output/IL15vsPC1.pdf", plot = p, device = "pdf", width = 5, height = 3.3)
```


### circulating IL15 change overtime -- figure 4L
```{r, fig.width=5, fig.height=3.3}
covid19.samples.pbmc <- covid19.samples[covid19.samples$material_type == "PBMC",] %>%
  mutate(sample_id = paste(subject_id, visit, sep = "_"))

IL15_alltime <- IL15 %>% 
  filter(subject_id %in% unique(PC1$subject_id)) %>%
  group_by(sample_id) %>% 
  dplyr::summarise_at(vars(test_value_log10), mean) %>%
  left_join(covid19.samples.pbmc, by = "sample_id") %>%
  filter(!is.na(days_from_symptom_onset_to_sample_drawn)) %>%
  left_join(PC1[,which(colnames(PC1) %in% c("subject_id", "PC1", "PC1class"))], by = "subject_id") %>%
  mutate(PC1class = factor(PC1class, levels = c("PC1_low","PC1_high")))

p <- ggplot(IL15_alltime)+
  geom_point(alpha=.4, shape=21,
             aes(x = days_from_symptom_onset_to_sample_drawn, y = test_value_log10, fill=PC1class), size = 3.5)+
  scale_fill_manual(name="Severity", values = c("deepskyblue1","red")) +
  geom_line(aes(x = days_from_symptom_onset_to_sample_drawn, y = test_value_log10, group = subject_id), alpha = 0.2)+
  stat_summary(aes(x = days_from_symptom_onset_to_sample_drawn, y = test_value_log10, group = PC1class), fun = median, alpha=0)+
  stat_smooth(aes(x = days_from_symptom_onset_to_sample_drawn, y = test_value_log10, group = PC1class, color=PC1class), se = TRUE, alpha=0.1, size = 1.5)+# group = 1 overwrite group id
  scale_color_manual(name="PC1 Class",values=c("deepskyblue1","red"))+
  geom_vline(xintercept = 20, color = "grey60", linetype='dashed')+
  theme(axis.text.x = element_text(angle = 90))+
  facet_wrap(~condition)+
  theme_bw()
p

# ggsave(filename = "output/IL15.TSO.pdf", plot = p, device = "pdf", width = 5, height = 3.3)

```


```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```
