---
title: "Effect size (NES) of DSM_high and DSM_low TSO association IFN plot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(openxlsx)
library(reshape2)
library(ggrepel)
```

## Effect size (NES) of DSM_high and DSM_low with TSO association respectively
#### using type-I IFN response as an example
**Generate Figure 3E**  

read in enrichment results tables and filter selected pathways for each celltype
Enrichment results from PC1/DSM group and continuous days_since_onset (TSO) interaction model reflecting the association with TSO in DSM high and DSM low groups respectively  

```{r input}
# filter selected pathways for each celltype
FGSEA_IN_PATH_HIGH <- "input/differential_expression/fgsea_tables/days_onset_in_PC1high"
FGSEA_IN_PATH_LOW <- "input/differential_expression/fgsea_tables/days_onset_in_PC1low"
FGSEA_IN_PATH_DIFF <- "input/differential_expression/fgsea_tables/diff_days_onset_PC1high-low"
remove_celltype_high <- paste(FGSEA_IN_PATH_HIGH, c("/Dblt", "/DPT", "/gated_out", "/PB_Plasmablasts", "/TissueResMemT", "/Unknown", "/Granulocytes"), "--model@PC1group_onsetcontinous_interaction--coef@days_onset_in_PC1high--fgsea.tsv", sep = "")
remove_celltype_low <- paste(FGSEA_IN_PATH_LOW, c("/Dblt", "/DPT", "/gated_out", "/PB_Plasmablasts", "/TissueResMemT", "/Unknown", "/Granulocytes"), "--model@PC1group_onsetcontinous_interaction--coef@days_onset_in_PC1low--fgsea.tsv", sep = "")
remove_celltype_diff <- paste(FGSEA_IN_PATH_DIFF, c("/Dblt", "/DPT", "/gated_out", "/PB_Plasmablasts", "/TissueResMemT", "/Unknown", "/Granulocytes"), "--model@PC1group_onsetcontinous_interaction--coef@diff_days_onset_PC1High-low--fgsea.tsv", sep = "")

files_high <- list.files(FGSEA_IN_PATH_HIGH, full.names = TRUE)
files_high <- files_high[-which(files_high %in% remove_celltype_high)]
files_low <- list.files(FGSEA_IN_PATH_LOW, full.names = TRUE)
files_low <- files_low[-which(files_low %in% remove_celltype_low)]
files_diff <- list.files(FGSEA_IN_PATH_DIFF, full.names = TRUE)
files_diff <- files_diff[-which(files_diff %in% remove_celltype_diff)]

fgsea_list_high <- lapply(files_high, function(x) read.delim(x, header = TRUE, sep = "\t"))
names(fgsea_list_high) <- gsub("--model@PC1group_onsetcontinous_interaction--coef@days_onset_in_PC1high--fgsea\\.tsv", "", basename(files_high))
fgsea_list_low <- lapply(files_low, function(x) read.delim(x, header = TRUE, sep = "\t"))
names(fgsea_list_low) <- gsub("--model@PC1group_onsetcontinous_interaction--coef@days_onset_in_PC1low--fgsea\\.tsv", "", basename(files_low))
fgsea_list_diff <- lapply(files_diff, function(x) read.delim(x, header = TRUE, sep = "\t"))
names(fgsea_list_diff) <- gsub("--model@PC1group_onsetcontinous_interaction--coef@diff_days_onset_PC1High-low--fgsea\\.tsv", "", basename(files_diff))


# read in selected pathways
pathways <- read.xlsx("input/selected_pathway_summary.xlsx", sheet = "reduced", startRow = 1)
fgsea_list_high_selected <- lapply(fgsea_list_high, function(x) {filter(x, pathway %in% pathways$Primary_gene_sets)})
fgsea_list_low_selected <- lapply(fgsea_list_low, function(x) {filter(x, pathway %in% pathways$Primary_gene_sets)})
fgsea_list_diff_selected <- lapply(fgsea_list_diff, function(x) {filter(x, pathway %in% pathways$Primary_gene_sets)})

fgsea_list_selected <- list()
for (i in names(fgsea_list_high_selected)) {
  fgsea_list_selected[[i]] <- bind_rows(fgsea_list_high_selected[[i]], fgsea_list_low_selected[[i]], .id = "PC1_group")
}

fgsea_selected_df <- bind_rows(fgsea_list_selected, .id = "celltype")
fgsea_diff_selected_df <- bind_rows(fgsea_list_diff_selected, .id = "celltype")

fgsea_selected_df$PC1_group <- replace(fgsea_selected_df$PC1_group, fgsea_selected_df$PC1_group == "1", "PC1_high")
fgsea_selected_df$PC1_group <- replace(fgsea_selected_df$PC1_group, fgsea_selected_df$PC1_group == "2", "PC1_low")
fgsea_selected_df$direction <- ifelse(fgsea_selected_df$NES>0, "up", "dn")
```
  
  
### plot IFN NES scatter
```{r, fig.width=6, fig.height=5}
fgsea_selected_df_IFN <- filter(fgsea_selected_df, pathway %in% c("GO_RESPONSE_TO_TYPE_I_INTERFERON"))
fgsea_diff_selected_df_IFN <- filter(fgsea_diff_selected_df, pathway %in% c("GO_RESPONSE_TO_TYPE_I_INTERFERON"))

fgsea_selected_df_IFN_wide <- dcast(fgsea_selected_df_IFN, celltype+pathway ~ PC1_group,value.var = "NES") %>%
  left_join(select(fgsea_diff_selected_df_IFN, "celltype", "padj"), by = "celltype") %>%
  mutate(diff_padj_sig = padj < 0.05)

p <- ggplot(fgsea_selected_df_IFN_wide, aes(x = PC1_low, y = PC1_high, label = celltype)) +
  geom_point(aes(color = diff_padj_sig), size = 3) +
  scale_color_manual(values=c("#495464", "#f05454")) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "blue")+
  geom_hline(yintercept = 0, linetype = "dashed", color = "#9ba4b4")+
  geom_vline(xintercept = 0, linetype = "dashed", color = "#9ba4b4")+
  xlim(-3.2,1.5)+
  ylim(-3.2,1.5)+
  theme_bw() +
  geom_text_repel(
    data = fgsea_selected_df_IFN_wide,
    aes(label = celltype),
    color = "#495464",
    size = 5,
    box.padding = unit(0.35, "lines"),
    point.padding = unit(0.35, "lines")
    )

p
# ggsave(filename = "output/TSO_PC1group_NES.pdf", plot = p, device = "pdf", width=6, height = 5)

```


```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```

