---
title: "Clinical-based clustering of COVID patients"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Cluster patients based on the following clinical parameters:

Oxygen: SpO2/FiO2 ratio

Inflammation: CRP, IL-6, TNF, IL-18 Neutrophil/lymphocyte ratio 

IFN-I: IP-10

coagulation/endothelial: D-dimer

T-cell related: IFNg, CXCL9 (downstream of IFNg), iL-17, IL-4, IL-13

Lymphopenia: lymphocyte counts (CBC)

Tissue damage: LDH 
```{r setup, include=FALSE}
data.folder <- "input.data"
output.folder <- "output/DSM"
dir.create(output.folder,recursive = T)
library(knitr)
library(ggplot2)
library(ggsci)
library(ggpubr)
library(ggfortify)
library(Hmisc)
library(GGally)
library(plsRglm)
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)


load(file.path(data.folder,"covid19.metadata.paper1.RData"),verbose = T)
```

### Feature and Sample Filtering
```{r clinical}
paramaters.of.interest <- c("SpO2/FiO2","CRP","IL-6","TNF-a","TNF-b","IL-18/IL-1f4","ANC/ALC Ratio","IP-10","D-Dimer","IFN-Gamma","CXCL9/MIG","CXCL9/ MIG",
                            "IL-17","IL-4","IL-13","Fibrinogen","Lymphocytes",#"Serology Spike","Serology Nucleocapsid"
                            "Lactate Dehydrogenase (LDH)","Platelets (PLT)")#,"Neutrophils")
batch.sample.lab.values.all <- subset(covid19.lab.results,subject_id %in% subset(covid19.samples,!is.na(batch))$subject_id 
                                  & tolower(test_name) %in% tolower(paramaters.of.interest) & test_id != "Lymphocytes_pct")

batch.sample.lab.values <- merge(batch.sample.lab.values.all,subset(covid19.samples,material_type == "PBMC"),#"Serum"),#
                                 by.x=c("subject_id","days_from_admission_to_test"),by.y=c("subject_id","days_from_admission_to_sample_drawn"),
                                 suffixes = c("",".duplicate"))
batch.sample.lab.values <- batch.sample.lab.values[,grep("duplicate$",colnames(batch.sample.lab.values),invert = T)]
batch.sample.lab.values.avg <- reshape2::dcast(batch.sample.lab.values,sample_bsi_id + subject_id + condition + sex + age + severity + 
                                                 outcome + ever_admitted_to_icu + 
                                                 days_from_symptom_onset_to_hospitalization + number_of_days_hospitalized + days_from_hospitalization_to_death +
                                                 sample_label + visit + days_from_symptom_onset_to_test + days_from_admission_to_test + batch + test_order ~ test_id,
                                               value.var = "test_value",fun.aggregate = function(x){mean(x,na.rm = T)})
batch.sample.lab.values.avg <- batch.sample.lab.values.avg[order(batch.sample.lab.values.avg$test_order),]

# use admission date if missing symptom onset
batch.sample.lab.values.avg[is.na(batch.sample.lab.values.avg$days_from_symptom_onset_to_test),]$days_from_symptom_onset_to_test <-
  batch.sample.lab.values.avg[is.na(batch.sample.lab.values.avg$days_from_symptom_onset_to_test),]$days_from_admission_to_test
```

### Clinical and cytokine parameters at first time point
```{r initial.timepoint,out.width="800px",fig.height=6}
batch.sample.lab.values.avg.T0 <- batch.sample.lab.values.avg[!duplicated(batch.sample.lab.values.avg$subject_id),]
batch.sample.lab.values.avg.T0$severity.outcome <- paste0(batch.sample.lab.values.avg.T0$severity,"-",batch.sample.lab.values.avg.T0$outcome)
severity.color <- ggsci::pal_jama()(4)
names(severity.color) <- sort(unique(batch.sample.lab.values.avg.T0$severity.outcome))

ggplot(batch.sample.lab.values.avg.T0,aes(visit,days_from_symptom_onset_to_test)) + geom_boxplot(outlier.shape = NA,width=0.5) + 
  geom_point(pch=21,size=3,alpha=0.75,position = position_jitter(width=0.2,h=0),aes(fill=severity.outcome)) + 
  theme_bw() + ggsci::scale_fill_jama() + ylab("Days from Symptom Onset")

# remove samples >= 30 days from symptom onset
max.days.since.syptom.onset <- 30
cat("Removing samples >=",max.days.since.syptom.onset,"-",
    subset(batch.sample.lab.values.avg.T0,days_from_symptom_onset_to_test >= max.days.since.syptom.onset)$subject_id)
batch.sample.lab.values.avg.T0 <- subset(batch.sample.lab.values.avg.T0,days_from_symptom_onset_to_test < max.days.since.syptom.onset)

ggplot(batch.sample.lab.values.avg.T0,aes(visit,days_from_admission_to_test)) + geom_boxplot(outlier.shape = NA,width=0.5) + 
  geom_point(pch=21,size=3,alpha=0.75,position = position_jitter(w=0.25,h=0),aes(fill=severity.outcome)) + 
  theme_bw() + ggsci::scale_fill_jama() + ylab("Days from Hospital Admission")

batch.sample.lab.values.avg.T0.meta <- batch.sample.lab.values.avg.T0[,setdiff(colnames(batch.sample.lab.values.avg.T0),unique(batch.sample.lab.values$test_id))]
rownames(batch.sample.lab.values.avg.T0.meta) <- batch.sample.lab.values.avg.T0.meta$subject_id
batch.sample.lab.values.avg.T0.meta$severity.outcome <- paste0(batch.sample.lab.values.avg.T0.meta$severity,"-",batch.sample.lab.values.avg.T0.meta$outcome)
batch.sample.lab.values.avg.T0 <- batch.sample.lab.values.avg.T0[,setdiff(colnames(batch.sample.lab.values.avg.T0),
                                                                          colnames(batch.sample.lab.values.avg.T0.meta))]

rownames(batch.sample.lab.values.avg.T0.meta) <- batch.sample.lab.values.avg.T0.meta$subject_id
rownames(batch.sample.lab.values.avg.T0) <- batch.sample.lab.values.avg.T0.meta$subject_id
colnames(batch.sample.lab.values.avg.T0) <- sapply(colnames(batch.sample.lab.values.avg.T0),function(x){unlist(strsplit(x,"_|\\.\\."))[1]})

# transform to log10
batch.sample.lab.values.avg.T0 <- log10(batch.sample.lab.values.avg.T0+0.01)

pheatmap::pheatmap(scale(batch.sample.lab.values.avg.T0),
                   annotation_row = batch.sample.lab.values.avg.T0.meta[,c("severity.outcome","age","sex",
                                                                           "days_from_symptom_onset_to_test","days_from_admission_to_test")],
                   color = viridis::inferno(101),fontsize = 6,
                   annotation_colors = list(sex=c("F"="darkred","M"="darkblue"),severity.outcome=severity.color),main="All earliest time-point batch samples")

# try to borrow from +/- 2 days
day.margin <- 2
batch.sample.lab.values.expanded <- data.frame()
for (i in 1:nrow(batch.sample.lab.values.avg.T0.meta)) {
  subj <- batch.sample.lab.values.avg.T0.meta$subject_id[i]
  #cat(i,"-",subj,"\n")
  days_from_admission <- batch.sample.lab.values.avg.T0.meta$days_from_admission_to_test[i]
  tmp <- subset(batch.sample.lab.values.all,subject_id == subj &
                  (days_from_admission_to_test == batch.sample.lab.values.avg.T0.meta$days_from_admission_to_test[i] |
    (days_from_admission_to_test <= (days_from_admission + day.margin) & 
      days_from_admission_to_test >= (days_from_admission - day.margin))))
  tmp$target.day.diff <- abs(tmp$days_from_admission_to_test - days_from_admission)
  #print(table(tmp$target.day.diff))
  
  tmp <- tmp[order(tmp$target.day.diff,decreasing = F),]
  tmp$days_from_admission_to_test_id <- apply(tmp[,c("subject_id","test_id","days_from_admission_to_test")],1,paste0,collapse=":")
  #tmp <- subset(tmp,days_from_symptom_onset_to_test_id %in% tmp[!duplicated(tmp[,c("subject_id","test_id")]),]$days_from_admission_to_test_id) # keep multiple values taken on the same closest day
  tmp <- tmp[!duplicated(tmp[,c("subject_id","test_id")],fromLast = F),]
  batch.sample.lab.values.expanded <- rbind(batch.sample.lab.values.expanded,tmp)
}

batch.sample.lab.values.expanded.avg <- reshape2::dcast(batch.sample.lab.values.expanded,subject_id ~ test_id,
                                               value.var = "test_value",fun.aggregate = function(x){mean(x,na.rm = T)})
rownames(batch.sample.lab.values.expanded.avg) <- batch.sample.lab.values.expanded.avg$subject_id
batch.sample.lab.values.expanded.avg <- batch.sample.lab.values.expanded.avg[rownames(batch.sample.lab.values.avg.T0.meta),
                                                                    intersect(colnames(batch.sample.lab.values.expanded.avg),batch.sample.lab.values.all$test_id)]


# transform to log10
batch.sample.lab.values.expanded.avg <- log10(batch.sample.lab.values.expanded.avg+0.01)

pheatmap::pheatmap(scale(batch.sample.lab.values.expanded.avg),
                   annotation_row = batch.sample.lab.values.avg.T0.meta[,c("severity.outcome","age","sex",
                                                                           "days_from_symptom_onset_to_test","days_from_admission_to_test")],
                   color = viridis::inferno(101),fontsize = 6,
                   annotation_colors = list(sex=c("F"="darkred","M"="darkblue"),severity.outcome=severity.color),
                   main="All earliest time-point batch samples (values from time point +/- 2 days)")

# impute remaining missing data
RNGkind(sample.kind = "Rounding") # ensure compatibility with versions prior to 3.6
#set.seed(10276)
set.seed(1087)
batch.sample.lab.values.expanded.avg.imputed <- batch.sample.lab.values.expanded.avg
imputed.obj <- aregImpute(formula = as.formula(paste0("~ ",paste0(colnames(batch.sample.lab.values.expanded.avg),collapse = " + "))), 
                                                           data = batch.sample.lab.values.expanded.avg, n.impute = 5, type = "regression",nk = 0)
imputed.values <- impute.transcan(imputed.obj,imputation = 1,data=batch.sample.lab.values.expanded.avg,list.out=TRUE)
batch.sample.lab.values.expanded.avg.imputed[names(imputed.values)] <- imputed.values

pheatmap::pheatmap(scale(batch.sample.lab.values.expanded.avg.imputed),
                   annotation_row = batch.sample.lab.values.avg.T0.meta[,c("severity.outcome","age","sex",
                                                                           "days_from_symptom_onset_to_test","days_from_admission_to_test")],
                   color = viridis::inferno(101),fontsize = 6,
                   annotation_colors = list(sex=c("F"="darkred","M"="darkblue"),severity.outcome=severity.color),
                   main="All earliest time-point batch samples (values from time point +/- 2 days or imputed)")

# correct for days from symptom onset or admission 
correct.for <- "days_from_admission_to_test"
batch.sample.lab.values.expanded.avg.imputed.corrected <- t(limma::removeBatchEffect(t(batch.sample.lab.values.expanded.avg.imputed),
                  covariates = batch.sample.lab.values.avg.T0.meta[rownames(batch.sample.lab.values.expanded.avg.imputed),correct.for,drop=F]))

parameter.labels <- unique(batch.sample.lab.values[,c("test_id","test_name")])
rownames(parameter.labels) <- parameter.labels$test_id
parameter.labels["ANC.ALC.Ratio_.ratio","test_name"] <- "Neutrophil/lymphocyte ratio (NLR)"
parameter.labels["CRP","test_name"] <- "C-reactive protein (CRP)"
parameter.labels["Lymphocytes_x10.3.uL","test_name"] <- "Lymphocyte count"
parameter.labels["Platelets..PLT._x10.3.uL","test_name"] <- "Platelet count"
parameter.labels$test_name <- gsub("/ ","/",parameter.labels$test_name,fixed = T)

# merge back patients with multiple samples
batch.sample.lab.values.expanded.avg.imputed.corrected <- 
  aggregate(batch.sample.lab.values.expanded.avg.imputed.corrected,list(batch.sample.lab.values.avg.T0.meta$subject_id),mean)
rownames(batch.sample.lab.values.expanded.avg.imputed.corrected) <- batch.sample.lab.values.expanded.avg.imputed.corrected$Group.1 
batch.sample.lab.values.expanded.avg.imputed.corrected <- batch.sample.lab.values.expanded.avg.imputed.corrected[,-1]
batch.sample.lab.values.avg.T0.meta <- batch.sample.lab.values.avg.T0.meta[match(rownames(batch.sample.lab.values.expanded.avg.imputed.corrected),
                                                                                 batch.sample.lab.values.avg.T0.meta$subject_id),]

# which were intubated?
batch.sample.lab.values.avg.T0.meta$intubated <- batch.sample.lab.values.avg.T0.meta$subject_id %in%  
  covid19.treatment[grep("int",covid19.treatment$treatment_notes,ignore.case = T),]$subject_id
subject.labels <- gsub("HGR0000","P",rownames(batch.sample.lab.values.expanded.avg.imputed.corrected))
subject.labels[batch.sample.lab.values.avg.T0.meta$ever_admitted_to_icu == T] <- 
  paste0(subject.labels[batch.sample.lab.values.avg.T0.meta$ever_admitted_to_icu == T],"*")
subject.labels[batch.sample.lab.values.avg.T0.meta$intubated == T] <- 
  paste0(subject.labels[batch.sample.lab.values.avg.T0.meta$intubated == T],"^")

pheatmap::pheatmap(scale(batch.sample.lab.values.expanded.avg.imputed.corrected),
                   annotation_row = batch.sample.lab.values.avg.T0.meta[,c("severity.outcome"),drop=F],#,"WHO.ordinal.scale.at.time.of.sampling"),drop=F],
                   #,"age","sex","batch","days_from_symptom_onset_to_test","days_from_admission_to_test")],
                   #color = colorRampPalette(brewer.pal(8,"YlGnBu"))(34),
                   color = viridis::cividis(34),
                   breaks = c(-3,seq(-1.5,1.5,by=0.1),3),angle_col = 45,cutree_rows = 4,
                   clustering_method = "ward.D",
                   annotation_colors = list(sex=c("F"="darkred","M"="darkblue"),severity.outcome=severity.color),
                   #main=paste0("Values from time point +/- 2 days or imputed corrected for ",correct.for),
                   labels_row = subject.labels,cellwidth = 15,cellheight = 8,fontsize = 8,#file=file.path(data.folder,"plots","test2.pdf"),
                   labels_col = parameter.labels[colnames(batch.sample.lab.values.expanded.avg.imputed.corrected),"test_name"])
```

# Classification model
Select features associated with severity and use them to calculate DSM
```{r model}
severity.index <- c("Moderate-Alive"=1,"Severe-Alive"=2,"Critical-Alive"=3,"Critical-Deceased"=4)
batch.sample.lab.values.expanded.avg.corrected <- t(limma::removeBatchEffect(t(batch.sample.lab.values.expanded.avg),
                  covariates = batch.sample.lab.values.avg.T0.meta[rownames(batch.sample.lab.values.expanded.avg),correct.for,drop=F]))
batch.sample.lab.values.expanded.avg.corrected <- batch.sample.lab.values.expanded.avg.corrected[rownames(batch.sample.lab.values.avg.T0.meta),
                                                                          -grep("FiO2",colnames(batch.sample.lab.values.expanded.avg.corrected))]

# leave-one-out modeling
feature.importance <- data.frame()
sample.scores <- data.frame()
for (i in 1:nrow(batch.sample.lab.values.expanded.avg.corrected)) {
  tryCatch({
    modpls2 <- plsRglm(#factor(who.severity.index[batch.sample.lab.values.avg.T0.meta$WHO.severity.outcome[-i]]),
                       factor(severity.index[batch.sample.lab.values.avg.T0.meta$severity.outcome[-i]]),
                       batch.sample.lab.values.expanded.avg.corrected[-i,],1,modele="pls-glm-polr",pvals.expli=T,
                       dataPredictY = batch.sample.lab.values.expanded.avg.corrected[i,,drop=F],verbose=F)
    sample.scores <- rbind(sample.scores,data.frame(sample=rownames(batch.sample.lab.values.expanded.avg.corrected)[i],
                                                    predicted=as.numeric(modpls2$ValsPredictYCat),
                                                    #observed=who.severity.index[batch.sample.lab.values.avg.T0.meta$WHO.severity.outcome[i]]))
                                                    observed=severity.index[batch.sample.lab.values.avg.T0.meta$severity.outcome[i]]))
    feature.importance <- rbind(feature.importance,as.data.frame(t(modpls2$Std.Coeffs)))

  }, error = function(e){warning(paste0("Error in sample",i,"\n"))})
}
cat("Contingency table:\n")
table(sample.scores[,c("predicted","observed")])
num.wrong <- sum(sample.scores$predicted != sample.scores$observed)
cat("Number of misclassified samples:",num.wrong," (",100*num.wrong/nrow(sample.scores),"%)\n")
ggplot(reshape::melt(feature.importance[,colnames(batch.sample.lab.values.expanded.avg.corrected)]),aes(variable,value)) + 
  geom_boxplot() + geom_hline(yintercept = 0,linetype="dashed",color="red") +
  theme_bw() + theme(axis.text.x = element_text(angle=90,hjust = 1,vjust = 0.5)) + ggtitle("ordinal logistic PLS leave-one-out models") + ylab("Coefficients")

# permutation-based feature selection
modpls2 <- plsRglm(factor(severity.index[batch.sample.lab.values.avg.T0.meta$severity.outcome]),
                   batch.sample.lab.values.expanded.avg.corrected,1,modele="pls-glm-polr",pvals.expli=T)
bootYT2 <- bootplsglm(modpls2, R=100, strata=unclass(factor(severity.index[batch.sample.lab.values.avg.T0.meta$severity.outcome])), sim="permutation")
plots.confints.bootpls(confints.bootpls(bootYT2,typeBCa=FALSE), legendpos = "topright",xaxisticks=FALSE)
boxplots.bootpls(bootYT2,xaxisticks=FALSE,ranget0=TRUE)

#saveRDS(feature.importance,file.path(data.folder,"end.point.PLS.coefficients.RDS"))
feature.median <- sort(abs(apply(feature.importance[,colnames(batch.sample.lab.values.expanded.avg.corrected)],2,median)))
feature.sign <- as.factor(sign(apply(feature.importance[,colnames(batch.sample.lab.values.expanded.avg.corrected)],2,median))[names(feature.median)])
feature.sd <- apply(feature.importance[,colnames(batch.sample.lab.values.expanded.avg.corrected)],2,sd)[names(feature.median)]
feature.median

names(feature.median) <- parameter.labels[names(feature.median),"test_name"]
ggplot(data.frame(feature=factor(names(feature.median),levels = names(feature.median)),abs.median.coef=feature.median,coef.sd=feature.sd,sign=feature.sign),
       aes(feature,-abs.median.coef*as.numeric(as.character(feature.sign)))) + geom_hline(yintercept = 0) + geom_bar(stat = "identity",fill="deepskyblue3") + 
  geom_errorbar(aes(ymin=-abs.median.coef*as.numeric(as.character(feature.sign))-coef.sd,
                ymax=-abs.median.coef*as.numeric(as.character(feature.sign))+coef.sd),width=.2,color="red") +
  scale_fill_manual(name="direction",values=c("darkblue","lightblue")) +
  theme_bw() + theme(axis.text.x=element_text(angle=45,hjust = 1,size=7)) + xlab("Parameter") + ylab("Median Coefficient") + ggtitle("PLS Model")

# final model
selected.num.features <- 8
# heatmap using important features + oxygen
important.features <- tail(names(feature.sign),selected.num.features)
print(important.features)

batch.sample.lab.values.avg.T0.meta$ever_admitted_to_icu <- as.character(batch.sample.lab.values.avg.T0.meta$ever_admitted_to_icu)
pheatmap::pheatmap(scale(batch.sample.lab.values.expanded.avg.imputed.corrected[,important.features]),
                   annotation_row = batch.sample.lab.values.avg.T0.meta[,c("severity.outcome","ever_admitted_to_icu","age","sex","batch",correct.for)],
                   color = viridis::inferno(34),breaks = c(-3,seq(-1.5,1.5,by=0.1),3),fontsize = 6,
                   annotation_colors = list(sex=c("F"="darkred","M"="darkblue"),severity.outcome=severity.color,
                                            ever_admitted_to_icu=c("FALSE"="white","TRUE"="black")),
                   cutree_rows = 4,
                   main=paste0("All earliest batch samples (values from time point +/- 2 days or imputed corrected for ",correct.for,")"))

# put oxygen back into PCA calculation
important.features <- c(important.features,"SpO2.FiO2_.ratio")
lab.pca <- prcomp(batch.sample.lab.values.expanded.avg.imputed.corrected[,important.features],scale. = T,center = T)
pca.summary <- summary(lab.pca)

row.annotation <- merge(batch.sample.lab.values.avg.T0.meta,lab.pca$x[,1:2],by="row.names")

# flip sign of PC1 to use it as "severity" metric
row.annotation$PC1 <- row.annotation$PC1*-1

# violin plot by severity group
comparisons <- gtools::combinations(4,2,names(severity.color))
comparisons <- comparisons[comparisons[,1] != comparisons[,2],]
comparisons <- lapply(1:nrow(comparisons),function(x){comparisons[x,]})
row.annotation$severity.outcome <- factor(row.annotation$severity.outcome,levels=names(severity.index))
ggviolin(row.annotation,x="severity.outcome",y="PC1", add = "boxplot",palette = c("#00A1D5FF","#B24745FF","#374E55FF","#DF8F44FF"),
         fill="severity.outcome",add.params = list(fill="white"),xlab = "Severity",ylab="Disease Severity Metric (DSM)") + 
  geom_point(alpha=0.25,size=2) + #scale_y_continuous(expand=expansion(add=c(0,1))) +
  stat_compare_means(comparisons = comparisons,hide.ns = T,label = "p.signif",tip.length = 0,method = "wilcox.test") + 
  theme(text = element_text(size=8)) 
rownames(row.annotation) <- row.annotation$Row.names
```


### Serology for batch samples
add serology data and create dsm object
```{r}
# serology at time point
serology.values <- subset(covid19.lab.results,test_type == "Serology")
serology.values$severity.outcome <- paste0(serology.values$severity,"-",serology.values$outcome)
serology.values.wide <- reshape2::dcast(serology.values,subject_id + days_from_symptom_onset_to_test + days_from_admission_to_test ~ test_name,value.var="test_value",
                                        fun.aggregate = mean)
colnames(serology.values.wide) <- gsub(" Antibodies","",colnames(serology.values.wide))
serology.values.wide$Nucleocapsid <- log10(serology.values.wide$Nucleocapsid+0.01)
serology.values.wide$Spike <- log10(serology.values.wide$Spike+0.01)
serology.values <- merge(covid19.patient.demo,serology.values.wide,by="subject_id",all.y=T)
colnames(serology.values) <- gsub("test","serology",colnames(serology.values))

# add individual parameters
patient.data <- row.annotation#cbind(row.annotation,batch.sample.lab.values.expanded.avg.imputed[row.annotation$subject_id,important.features])
batch.serology <- subset(serology.values,subject_id %in% patient.data$subject_id)

# serology closest to PBMC time point
batch.serology$sample.time.point <- patient.data[batch.serology$subject_id,]$days_from_admission_to_test
batch.serology$days_from_sample_to_serology <- batch.serology$days_from_admission_to_serology - batch.serology$sample.time.point
batch.serology <- batch.serology[order(abs(batch.serology$days_from_sample_to_serology)),]
batch.serology.time.point <- batch.serology[!duplicated(batch.serology$subject_id),]
patient.data <- cbind(patient.data,initial=batch.serology.time.point[match(patient.data$subject_id,batch.serology.time.point$subject_id),
                                                  c("Nucleocapsid","Spike","days_from_sample_to_serology","days_from_admission_to_serology")])

# serology peak titer
batch.serology <- batch.serology[order(batch.serology$Nucleocapsid,decreasing = T),]
batch.serology$peak.Nucleocapsid <- !duplicated(batch.serology$subject_id)
Nucleocapsid.peak.titer <- batch.serology[!duplicated(batch.serology$subject_id),]
patient.data <- cbind(patient.data,nucleocapsid.peak=Nucleocapsid.peak.titer[match(patient.data$subject_id,Nucleocapsid.peak.titer$subject_id),
                                                  c("Nucleocapsid","Spike","days_from_sample_to_serology","days_from_admission_to_serology")])

batch.serology <- batch.serology[order(batch.serology$Spike,decreasing = T),]
batch.serology$peak.Spike <- !duplicated(batch.serology$subject_id)
Spike.peak.titer <- batch.serology[!duplicated(batch.serology$subject_id),]
patient.data <- cbind(patient.data,spike.peak=Spike.peak.titer[match(patient.data$subject_id,Spike.peak.titer$subject_id),
                                                  c("Nucleocapsid","Spike","days_from_sample_to_serology","days_from_admission_to_serology")])

# serology closest to discharge or death
batch.serology$days_from_hospitalization_end_to_serology <- as.numeric(batch.serology$days_from_admission_to_serology - batch.serology$number_of_days_hospitalized)
batch.serology[which(is.na(batch.serology$days_from_hospitalization_end_to_serology)),]$days_from_hospitalization_end_to_serology <- 
  as.numeric(batch.serology$days_from_admission_to_serology - 
               batch.serology$days_from_hospitalization_to_death)[is.na(batch.serology$days_from_hospitalization_end_to_serology)]
batch.serology <- batch.serology[order(abs(batch.serology$days_from_hospitalization_end_to_serology)),]

end.batch.serology <- batch.serology[!duplicated(batch.serology$subject_id),]
patient.data <- cbind(patient.data,end=end.batch.serology[match(patient.data$subject_id,end.batch.serology$subject_id),
                                                  c("Nucleocapsid","Spike","days_from_hospitalization_end_to_serology","days_from_admission_to_serology")])

```

### Severity by Sample
WHO serverity score over time between DSM-high and DSM-low groups
```{r WHO.score}
who.score.df <- read.csv(file.path(data.folder,"pbmc.samples.csv"))
who.score.df <- merge(who.score.df,subset(covid19.samples,material_type == "PBMC"),by = c("subject_id","sample_label","visit"),suffixes = c("",".dup"))
who.score.df$DSM.group <- NA
who.score.df[who.score.df$subject_id %in% subset(patient.data,PC1 < median(patient.data$PC1))$subject_id,"DSM.group"] <- "DSM-low"
who.score.df[who.score.df$subject_id %in% subset(patient.data,PC1 > median(patient.data$PC1))$subject_id,"DSM.group"] <- "DSM-high"

ggplot(subset(who.score.df,!is.na(DSM.group)),aes(days_from_symptom_onset_to_sample_drawn,WHO.ordinal.scale.at.time.of.sampling)) + 
  geom_rect(aes(xmin=17,xmax=23,ymax=Inf,ymin=-Inf),fill="lightyellow2",alpha=0.25) + geom_point(aes(fill=DSM.group),pch=21,size=3,alpha=0.25) + 
  geom_line(aes(group=subject_id),alpha=0.1) + theme_bw() + 
  xlab("TSO") + ylab("WHO Sample Severity Score") + geom_smooth(se = T,aes(color=DSM.group),alpha=0.15,size=1.25) + 
  scale_color_manual(name="DSM group",values=c("red","deepskyblue1")) + scale_fill_manual(name="DSM group",values=c("red","deepskyblue1"))

# DSM v. WHO Scores
dsm.who.association <- merge(patient.data,who.score.df[,c("sample_label","visit","WHO.ordinal.scale.at.time.of.sampling")],
                                by=c("sample_label","visit"),all.x=T)
ggplot(dsm.who.association,aes(as.factor(WHO.ordinal.scale.at.time.of.sampling),PC1)) + geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha=0.75,size=2,pch=21,aes(fill=severity.outcome),width=0.1,height=0) + xlab("WHO Severity Score") + ylab("DSM") +
  stat_compare_means() +
  scale_fill_manual(name="Severity-Outcome",values=severity.color) + theme_bw() + ylim(-3,4.5)

dsm.who.association$WHO.severity <- "Mild"
dsm.who.association[dsm.who.association$WHO.ordinal.scale.at.time.of.sampling >= 5,"WHO.severity"] <- "Severe"
ggplot(dsm.who.association,aes(as.factor(WHO.severity),PC1)) + geom_boxplot(outlier.shape = NA) +
geom_jitter(alpha=0.75,size=2,pch=21,aes(fill=severity.outcome),width=0.1,height=0) + xlab("WHO Severity Class") + ylab("DSM") +
    stat_compare_means() +
  scale_fill_manual(name="Severity-Outcome",values=severity.color) + theme_bw() + ylim(-3,4.5)
```

### Session Info
```{r save}
patient.data <- patient.data[,-c(1:2,18:20)]
colnames(patient.data) <- gsub("test","sample",colnames(patient.data))
patient.data[is.na(patient.data$date_of_onset_of_symptom),"days_from_symptom_onset_to_sample"] <- NA
patient.data <- patient.data[,grep("date",colnames(patient.data),invert = T)]
saveRDS(patient.data,file.path(output.folder,"citeseq.patient.end.points.RDS"))

sessionInfo()
```
