---
title: "Classification of Brescia Deceased vs. Recovered"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Compare cytokine profiles between deceased patients and those with good outcome. Specific focus on 17-23 days since symptom onset.
```{r setup, include=FALSE}
data.folder <- "input.data"
output.folder <- "output"
library(knitr)
library(ggplot2)
library(ggsci)
library(ggpubr)
library(ggfortify)
library(mgcv)
library(tidymv)
library(lme4)
library(lmerTest)
library(plsRglm)
library(precrec)
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)

source("critical.juncture/juncture.protein.search.R")

# parallel processing for permutation test
library(BiocParallel)
library(parallel)
library(foreach)
num.workers <- 60
if (!exists("cl")) {
  register(MulticoreParam(num.workers, progressbar = FALSE))
  cl <- parallel::makeForkCluster(num.workers,outfile="cluster.log.txt")
  doParallel::registerDoParallel(cl)
  cat("Number of workers started: ",getDoParWorkers(),"\n")
}

if (as.numeric(paste0(R.version$major,".",floor(as.numeric(R.version$minor)))) >= 3.6)
  RNGkind(sample.kind = "Rounding") # ensure compatibility with versions prior to 3.6
```

### Create sub-cohort
Identify all deceased patients and match them with critical but recovered patients by age and gender (excluding those staying hospitalized for more than 50 days).
Patients in the citiseq cohort can optionally be excluded.
```{r read.data}
load(file.path(data.folder,"covid19.metadata.paper1.RData"),verbose=F)
covid19.lab.results$severity.outcome <- paste0(covid19.lab.results$severity,"-",covid19.lab.results$outcome)

severity.color <- c("Critical-Alive"="#374E55FF","Critical-Deceased"="#DF8F44FF","Moderate-Alive"="#00A1D5FF" ,"Severe-Alive"="#B24745FF","HC" = "#79AF97FF")

deceased.patients.cytokines <- subset(covid19.lab.results,outcome == "Deceased" & test_type == "ELISA" & !is.na(days_from_symptom_onset_to_test))
deceased.patients <- unique(deceased.patients.cytokines$subject_id)

cat("Map of sampling TSO time points for deceased patients:\n")
deceased.time.points <- as.data.frame.matrix(table(deceased.patients.cytokines[,c("subject_id","days_from_symptom_onset_to_test")])) 
colSums(deceased.time.points > 0)
cat("Age and sex distribution:\n")
table(subset(covid19.patient.demo,subject_id %in% deceased.patients)$sex)
summary(subset(covid19.patient.demo,subject_id %in% deceased.patients)$age)

# recovered patients with similar age and sex
age.tol <- 3
recovered.patients <- c()
for (i in deceased.patients) {
  target.age <- covid19.patient.demo[i,"age"]
  target.sex <- covid19.patient.demo[i,"sex"]
  cat("Target age:",target.age,"sex:",target.sex,"\n")
  tmp <- subset(covid19.lab.results,outcome != "Deceased" & test_type == "ELISA" & !is.na(days_from_symptom_onset_to_test) & 
           age >= target.age - age.tol & age <= target.age + age.tol & severity == "Critical" & sex == target.sex & 
             days_from_symptom_onset_to_test > 15 & days_from_symptom_onset_to_test <= 25 &
           number_of_days_hospitalized < 50)

  matched.subject <- unique(tmp$subject_id)
  cat("Matched subject:",matched.subject,"\n")
  recovered.patients <- c(recovered.patients,matched.subject)
}
recovered.patients <- unique(recovered.patients[!is.na(recovered.patients)])
recovered.patients.cytokines <- subset(covid19.lab.results,test_type == "ELISA" & subject_id %in% recovered.patients)
cat("Map of samping TSO time points for recovered patients:\n")
recovered.time.points <- as.data.frame.matrix(table(recovered.patients.cytokines[,c("subject_id","days_from_symptom_onset_to_test")])) 
colSums(recovered.time.points > 0)
cat("Age and sex distribution:\n")
table(subset(covid19.patient.demo,subject_id %in% recovered.patients)$sex)
summary(subset(covid19.patient.demo,subject_id %in% recovered.patients)$age)

selected.cytokine.profiles <- rbind(deceased.patients.cytokines,recovered.patients.cytokines)
# remove patients in citeseq
#citeseq.patients <- subset(covid19.samples,!is.na(batch))$subject_id
#intersect(citeseq.patients,c(recovered.patients,deceased.patients))
#selected.cytokine.profiles <- subset(selected.cytokine.profiles,!(subject_id %in% citeseq.patients)) 
parameter.labels <- unique(selected.cytokine.profiles[,c("test_id","test_name")])
rownames(parameter.labels) <- parameter.labels$test_id
parameter.labels$test_name <- gsub("/ ","/",parameter.labels$test_name,fixed = T)
```


### Grouping of samples
Separate samples into interval groups relative to juncture: early (7-16 days), mid (juncture/17-23 days), and late (24-30 days)
```{r build}
selected.cytokine.profiles$unique.id <- make.unique(apply(selected.cytokine.profiles[,c("subject_id","days_from_admission_to_test","test_id")],1,
                                                          paste0,collapse="."),sep = "|")
selected.cytokine.profiles$unique.id <- sapply(selected.cytokine.profiles$unique.id,function(x){unlist(strsplit(x,"|",fixed=T))[2]})
selected.cytokine.profiles.matrix <- reshape2::dcast(selected.cytokine.profiles,
                                                     subject_id + age + sex + outcome + severity + days_from_symptom_onset_to_test + unique.id ~ test_id,
                                                     value.var = "test_value")
selected.cytokine.profiles.matrix[,colnames(selected.cytokine.profiles.matrix) %in% selected.cytokine.profiles$test_id] <-
  apply(selected.cytokine.profiles.matrix[,colnames(selected.cytokine.profiles.matrix) %in% selected.cytokine.profiles$test_id],2,function(x){scale(log10(x+1))})
selected.cytokine.profiles.matrix$stage <- NA
tmp <- which(selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test >= 7 & selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test <= 16)
selected.cytokine.profiles.matrix[tmp,"stage"] <- "early"
tmp <- which(selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test >= 17 & selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test <= 23)
selected.cytokine.profiles.matrix[tmp,"stage"] <- "mid"
tmp <- which(selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test >= 24 & selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test <= 30)
selected.cytokine.profiles.matrix[tmp,"stage"] <- "late"

cat("Sample distribution:\n")
table(selected.cytokine.profiles.matrix[,c("outcome","stage")])
```
<br/>
    
### Outcome Prediction 
Search for divergent proteins around juncture using the following model:

Differential expression: 
test_value ~ outcome + age.numeric + sex + DFSO + (1|subject_id), for samples within each stage to identify cytokines
with sig. differences between the two outcome groups

Do leave-one-out cross validation by holding out one patient in each round for testing
```{r predict}
# leave-one-out modeling
sample.scores <- data.frame()
feature.importance <- data.frame()
selected.cytokine.profiles.matrix$outcome.numerical <- 0
selected.cytokine.profiles.matrix[selected.cytokine.profiles.matrix$outcome == "Deceased",]$outcome.numerical <- 1
for (i in unique(selected.cytokine.profiles.matrix$subject_id)) {
  train.data <- subset(selected.cytokine.profiles.matrix,subject_id != i)
  test.data <- subset(selected.cytokine.profiles.matrix,subject_id == i)
  
  # feature selection
  search.results <- juncture.divergent.proteins(train.data,pval.cutoff = 1)

  # modeling
  for (s in c("mid","late")) {
    pls.train.data <- subset(train.data,stage == s)
    pls.test.data <- subset(test.data,stage == s)
    sig.did.de.df <- subset(search.results$type1,de.pval.1 <= 0.05 & 
                                 did.pval.test.intervalTRUE.outcomeDeceased <= 0.05 & stage == s) # DE & sig. delta 
    if (nrow(pls.test.data) > 0 & nrow(pls.train.data) > 0) {
      modpls2 <-plsRglm(dataY=pls.train.data$outcome.numerical,dataX=pls.train.data[,sig.did.de.df$cytokine],
                        nt=1,modele="pls-glm-logistic",pvals.expli = T,
                      dataPredictY = pls.test.data[,sig.did.de.df$cytokine,drop=F],sparse = F,verbose = F)
      sample.scores <- rbind(sample.scores,data.frame(stage=s,sample=pls.test.data[,
                                                    c("subject_id","days_from_symptom_onset_to_test","stage","outcome")],
                                                    predicted=modpls2$ValsPredictY,
                                                    observed=pls.test.data[,"outcome.numerical"]))
      feature.importance <- dplyr::bind_rows(feature.importance,data.frame(stage=s,t(modpls2$Std.Coeffs)))
    }
  }
}
# performance
cat("Performance at juncture:\n")
precrec_obj_mid <- evalmod(scores = subset(sample.scores,stage=="mid")$predicted,labels = subset(sample.scores,stage=="mid")$observed)
precrec_obj_mid
cat("Performance post-juncture:\n")
precrec_obj_late <- evalmod(scores = subset(sample.scores,stage=="late")$predicted,labels = subset(sample.scores,stage=="late")$observed)
precrec_obj_late
#precrec_obj <- evalmod(scores = sample.scores$predicted,labels = sample.scores$observed)
#autoplot(precrec_obj)

# ROC plot
roc.curve.all <- rbind(data.frame(type="mid",x=precrec_obj_mid$rocs[[1]]$x,y=precrec_obj_mid$rocs[[1]]$y),
                       data.frame(type="late",x=precrec_obj_late$rocs[[1]]$x,y=precrec_obj_late$rocs[[1]]$y))
ggplot(roc.curve.all,aes(x,y,color=type)) + geom_path(size=1.5,alpha=0.95) + theme_bw() +
 geom_abline(slope=1,intercept = 0,linetype="dashed",color="grey30") +
 xlab("1-Specificity") + ylab("Sensitivity") + scale_color_manual(labels=c("Post-Juncture","Juncture"),values=c("lightblue","#EADCFA"))
```

### Permutation
Calculate permutation-based p-value of classification performance by permuting the patient labels
```{r permute, eval=F}
# permute labels
num.perm <- 1000
protein.search.res.permuted <- mclapply(which(sapply(permuted.auc,is.null)==T),function(j) {
  set.seed(j)
  cat("Round",j,"\n")
  selected.cytokine.profiles.permute <- selected.cytokine.profiles.matrix
  subject.outcome <- unique(selected.cytokine.profiles.permute[,c("subject_id","outcome")])
  subject.outcome$permuted.outcome <- sample(subject.outcome$outcome,nrow(subject.outcome))
  selected.cytokine.profiles.permute$outcome <- subject.outcome[match(selected.cytokine.profiles.permute$subject_id,subject.outcome$subject_id),
                                                                "permuted.outcome"]
  #return(selected.cytokine.profiles.permute)
  
  # leave-one-out modeling
  sample.scores <- data.frame()
  feature.importance <- data.frame()
  selected.cytokine.profiles.permute$outcome.numerical <- 0
  selected.cytokine.profiles.permute[selected.cytokine.profiles.permute$outcome == "Deceased",]$outcome.numerical <- 1
  for (i in unique(selected.cytokine.profiles.matrix$subject_id)) {
    cat(i,"\n")
    train.data <- subset(selected.cytokine.profiles.permute,subject_id != i)
    test.data <- subset(selected.cytokine.profiles.permute,subject_id == i)
    
    # feature selection
    search.results <- juncture.divergent.proteins(train.data,pval.cutoff = 1)
    
    # modeling
    for (s in c("mid","late")) {
      pls.train.data <- subset(train.data,stage == s)
      pls.test.data <- subset(test.data,stage == s)
      # feature selection
      sig.did.de.df <- subset(search.results$type1,de.pval.1 <= 0.05 & 
                                   did.pval.test.intervalTRUE.outcomeDeceased <= 0.05 & stage == s) # DE & sig. delta 
      if (nrow(pls.test.data) > 0 & nrow(pls.train.data) > 0 & nrow(sig.did.de.df) > 0) {
        modpls2 <-plsRglm(dataY=pls.train.data$outcome.numerical,dataX=pls.train.data[,sig.did.de.df$cytokine,drop=F],
                          nt=1,modele="pls-glm-logistic",pvals.expli = T,
                          dataPredictY = pls.test.data[,sig.did.de.df$cytokine,drop=F],sparse = F,verbose = T)
        if (is.numeric(modpls2$ValsPredictY)) {
          sample.scores <- rbind(sample.scores,data.frame(round=j,stage=s,sample=pls.test.data[,
                                                        c("subject_id","days_from_symptom_onset_to_test","stage","outcome")],
                                                        predicted=modpls2$ValsPredictY,
                                                        observed=pls.test.data[,"outcome.numerical"]))
          feature.importance <- dplyr::bind_rows(feature.importance,data.frame(round=j,stage=s,t(modpls2$Std.Coeffs)))
        }
      }
    }
  }
  res <- list(scores=sample.scores,features=feature.importance)
  return(res)
},mc.cores=num.workers)

permuted.auc <- lapply(1:length(protein.search.res.permuted),function(x){
  cat(x,"\n")
  if (class(protein.search.res.permuted[[x]]) != "try-error") {
    permuted.scores <- protein.search.res.permuted[[x]]$scores
    roc.auc <- c("mid"=NA,"late"=NA)
    for (s in unique(permuted.scores$stage)) {
      true.scores <- subset(sample.scores,stage == s)
      round.scores <- subset(permuted.scores,stage == s)
      true.scores <- merge(true.scores,round.scores[,c("sample.subject_id","sample.days_from_symptom_onset_to_test","predicted","observed")],
                           by = c("sample.subject_id","sample.days_from_symptom_onset_to_test"),all.x=T,suffixes=c("",".perm"))
      # default to score of 0.5 when no model is created
      true.scores[is.na(true.scores$predicted.perm),"predicted.perm"] <- 0.5
      auc <- evalmod(scores = true.scores$predicted.perm,labels = true.scores$observed)
      roc.auc[s] <- attr(auc$rocs[[1]],"auc")
    }
    return(roc.auc)
  } else {
    cat("error!\n")
  }
})

permuted.mid.auc <- sapply(permuted.auc,function(x){x["mid"]})
cat("P-value for juncture samples:",sum(permuted.mid.auc > attr(precrec_obj_mid$rocs[[1]],"auc"),na.rm = T)/num.perm)
permuted.late.auc <- sapply(permuted.auc,function(x){x["late"]})
cat("P-value for post-juncture samples:",sum(permuted.late.auc > attr(precrec_obj_late$rocs[[1]],"auc"),na.rm = T)/num.perm)

```

### Session Info
```{r save}
stopCluster(cl)
sessionInfo()
```
