---
title: "Permuted Cytokine Profiles for Brescia Deceased vs. Recovered"
author: "William Lau"
date: "Dec 14, 2020"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
data.folder <- "input.data"
output.folder <- "output"
library(knitr)
library(ggplot2)
library(ggsci)
library(ggpubr)
library(ggfortify)
library(mgcv)
library(tidymv)
library(lme4)
library(lmerTest)
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)

# parallel processing for permutation test
library(BiocParallel)
library(parallel)
library(foreach)
num.workers <- 60
if (!exists("cl")) {
  register(MulticoreParam(num.workers, progressbar = FALSE))
  cl <- parallel::makeForkCluster(num.workers,outfile="cluster.log.txt")
  doParallel::registerDoParallel(cl)
  cat("Number of workers started: ",getDoParWorkers(),"\n")
}

source("critical.juncture/juncture.protein.search.R")
```

### Create sub-cohort
Identify all deceased patients and match them with critical but recovered patients by age and gender (excluding those staying hospitalized for more than 50 days).
Patients in the citiseq cohort can optionally be excluded.
```{r read.data}
load(file.path(data.folder,"covid19.metadata.paper1.RData"),verbose=F)
covid19.lab.results$severity.outcome <- paste0(covid19.lab.results$severity,"-",covid19.lab.results$outcome)

severity.color <- c("Critical-Alive"="#374E55FF","Critical-Deceased"="#DF8F44FF","Moderate-Alive"="#00A1D5FF" ,"Severe-Alive"="#B24745FF","HC" = "#79AF97FF")

deceased.patients.cytokines <- subset(covid19.lab.results,outcome == "Deceased" & test_type == "ELISA" & !is.na(days_from_symptom_onset_to_test))
deceased.patients <- unique(deceased.patients.cytokines$subject_id)

cat("Map of sampling TSO time points for deceased patients:\n")
deceased.time.points <- as.data.frame.matrix(table(deceased.patients.cytokines[,c("subject_id","days_from_symptom_onset_to_test")])) 
colSums(deceased.time.points > 0)
cat("Age and sex distribution:\n")
table(subset(covid19.patient.demo,subject_id %in% deceased.patients)$sex)
summary(subset(covid19.patient.demo,subject_id %in% deceased.patients)$age)

# recovered patients with similar age and sex
age.tol <- 3
recovered.patients <- c()
for (i in deceased.patients) {
  target.age <- covid19.patient.demo[i,"age"]
  target.sex <- covid19.patient.demo[i,"sex"]
  cat("Target age:",target.age,"sex:",target.sex,"\n")
  tmp <- subset(covid19.lab.results,outcome != "Deceased" & test_type == "ELISA" & !is.na(days_from_symptom_onset_to_test) & 
           age >= target.age - age.tol & age <= target.age + age.tol & severity == "Critical" & sex == target.sex & 
             days_from_symptom_onset_to_test > 15 & days_from_symptom_onset_to_test <= 25 &
           number_of_days_hospitalized < 50)

  matched.subject <- unique(tmp$subject_id)
  cat("Matched subject:",matched.subject,"\n")
  recovered.patients <- c(recovered.patients,matched.subject)
}
recovered.patients <- unique(recovered.patients[!is.na(recovered.patients)])
recovered.patients.cytokines <- subset(covid19.lab.results,test_type == "ELISA" & subject_id %in% recovered.patients)
cat("Map of samping TSO time points for recovered patients:\n")
recovered.time.points <- as.data.frame.matrix(table(recovered.patients.cytokines[,c("subject_id","days_from_symptom_onset_to_test")])) 
colSums(recovered.time.points > 0)
cat("Age and sex distribution:\n")
table(subset(covid19.patient.demo,subject_id %in% recovered.patients)$sex)
summary(subset(covid19.patient.demo,subject_id %in% recovered.patients)$age)

selected.cytokine.profiles <- rbind(deceased.patients.cytokines,recovered.patients.cytokines)
# remove patients in citeseq
#citeseq.patients <- subset(covid19.samples,!is.na(batch))$subject_id
#intersect(citeseq.patients,c(recovered.patients,deceased.patients))
#selected.cytokine.profiles <- subset(selected.cytokine.profiles,!(subject_id %in% citeseq.patients)) 
parameter.labels <- unique(selected.cytokine.profiles[,c("test_id","test_name")])
rownames(parameter.labels) <- parameter.labels$test_id
parameter.labels$test_name <- gsub("/ ","/",parameter.labels$test_name,fixed = T)
```


### Grouping of samples
Separate samples into interval groups relative to juncture: early (7-16 days), mid (juncture/17-23 days), and late (24-30 days)
```{r build}
selected.cytokine.profiles$unique.id <- make.unique(apply(selected.cytokine.profiles[,c("subject_id","days_from_admission_to_test","test_id")],1,
                                                          paste0,collapse="."),sep = "|")
selected.cytokine.profiles$unique.id <- sapply(selected.cytokine.profiles$unique.id,function(x){unlist(strsplit(x,"|",fixed=T))[2]})
selected.cytokine.profiles.matrix <- reshape2::dcast(selected.cytokine.profiles,
                                                     subject_id + age + sex + outcome + severity + days_from_symptom_onset_to_test + unique.id ~ test_id,
                                                     value.var = "test_value")
selected.cytokine.profiles.matrix[,colnames(selected.cytokine.profiles.matrix) %in% selected.cytokine.profiles$test_id] <-
  apply(selected.cytokine.profiles.matrix[,colnames(selected.cytokine.profiles.matrix) %in% selected.cytokine.profiles$test_id],2,function(x){scale(log10(x+1))})
selected.cytokine.profiles.matrix$stage <- NA
tmp <- which(selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test >= 7 & selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test <= 16)
selected.cytokine.profiles.matrix[tmp,"stage"] <- "early"
tmp <- which(selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test >= 17 & selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test <= 23)
selected.cytokine.profiles.matrix[tmp,"stage"] <- "mid"
tmp <- which(selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test >= 24 & selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test <= 30)
selected.cytokine.profiles.matrix[tmp,"stage"] <- "late"

cat("Sample distribution:\n")
table(selected.cytokine.profiles.matrix[,c("outcome","stage")])
```
<br/>


### Permutation
Calculate protein search FDR by permutation of subject outcome labels
```{r permute.search}
# permute labels
num.perm <- 1000
protein.search.res.permuted <- mclapply(1:num.perm,function(j) {
  set.seed(j)
  cat("Round",j,"\n")
  selected.cytokine.profiles.permute <- selected.cytokine.profiles.matrix
  subject.outcome <- unique(selected.cytokine.profiles.permute[,c("subject_id","outcome")])
  subject.outcome$permuted.outcome <- sample(subject.outcome$outcome,nrow(subject.outcome))
  selected.cytokine.profiles.permute$outcome <- subject.outcome[match(selected.cytokine.profiles.permute$subject_id,subject.outcome$subject_id),
                                                                "permuted.outcome"]
  
  res <- juncture.divergent.proteins(selected.cytokine.profiles.permute)
  return(res)
},mc.cores=num.workers)

type1.protein.freq <- do.call("rbind",lapply(protein.search.res.permuted,function(x){x$type1}))
cat("Protein permutation p-value:\n")
table(type1.protein.freq$cytokine)/num.perm

```

### Session Info
```{r save}
parallel::stopCluster(cl)
# save all results
dir.create(file.path(output.folder,"critical.juncture"))
save(protein.search.res.permuted,
     file = file.path(output.folder,"critical.juncture","deceased.v.recovered.cytokine.test.permutation.results.RData"))

sessionInfo()
```
