---
title: "Cytokine Profiles for Brescia Deceased vs. Recovered"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Compare cytokine profiles between deceased patients and those with good outcome. Specific focus on 17-23 days since symptom onset.
```{r setup, include=FALSE}
data.folder <- "input.data"
output.folder <- "output/critical.juncture"
dir.create(output.folder,recursive = T)
library(knitr)
library(ggplot2)
library(ggsci)
library(ggpubr)
library(ggfortify)
library(mgcv)
library(tidymv)
library(lme4)
library(lmerTest)
library(plsRglm)
library(precrec)
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)

source("critical.juncture/juncture.protein.search.R")

```

### Create sub-cohort
Identify all deceased patients and match them with critical but recovered patients by age and gender (excluding those staying hospitalized for more than 50 days).
Patients in the citiseq cohort can optionally be excluded.
```{r read.data}
load(file.path(data.folder,"covid19.metadata.paper1.RData"),verbose=F)
covid19.lab.results$severity.outcome <- paste0(covid19.lab.results$severity,"-",covid19.lab.results$outcome)

severity.color <- c("Critical-Alive"="#374E55FF","Critical-Deceased"="#DF8F44FF","Moderate-Alive"="#00A1D5FF" ,"Severe-Alive"="#B24745FF","HC" = "#79AF97FF")

deceased.patients.cytokines <- subset(covid19.lab.results,outcome == "Deceased" & test_type == "ELISA" & !is.na(days_from_symptom_onset_to_test))
deceased.patients <- unique(deceased.patients.cytokines$subject_id)

cat("Map of sampling TSO time points for deceased patients:\n")
deceased.time.points <- as.data.frame.matrix(table(deceased.patients.cytokines[,c("subject_id","days_from_symptom_onset_to_test")])) 
colSums(deceased.time.points > 0)
cat("Age and sex distribution:\n")
table(subset(covid19.patient.demo,subject_id %in% deceased.patients)$sex)
summary(subset(covid19.patient.demo,subject_id %in% deceased.patients)$age)

# recovered patients with similar age and sex
age.tol <- 3
recovered.patients <- c()
for (i in deceased.patients) {
  target.age <- covid19.patient.demo[i,"age"]
  target.sex <- covid19.patient.demo[i,"sex"]
  cat("Target age:",target.age,"sex:",target.sex,"\n")
  tmp <- subset(covid19.lab.results,outcome != "Deceased" & test_type == "ELISA" & !is.na(days_from_symptom_onset_to_test) & 
           age >= target.age - age.tol & age <= target.age + age.tol & severity == "Critical" & sex == target.sex & 
             days_from_symptom_onset_to_test > 15 & days_from_symptom_onset_to_test <= 25 &
           number_of_days_hospitalized < 50)

  matched.subject <- unique(tmp$subject_id)
  cat("Matched subject:",matched.subject,"\n")
  recovered.patients <- c(recovered.patients,matched.subject)
}
recovered.patients <- unique(recovered.patients[!is.na(recovered.patients)])
recovered.patients.cytokines <- subset(covid19.lab.results,test_type == "ELISA" & subject_id %in% recovered.patients)
cat("Map of samping TSO time points for recovered patients:\n")
recovered.time.points <- as.data.frame.matrix(table(recovered.patients.cytokines[,c("subject_id","days_from_symptom_onset_to_test")])) 
colSums(recovered.time.points > 0)
cat("Age and sex distribution:\n")
table(subset(covid19.patient.demo,subject_id %in% recovered.patients)$sex)
summary(subset(covid19.patient.demo,subject_id %in% recovered.patients)$age)

selected.cytokine.profiles <- rbind(deceased.patients.cytokines,recovered.patients.cytokines)
# remove patients in citeseq
#citeseq.patients <- subset(covid19.samples,!is.na(batch))$subject_id
#intersect(citeseq.patients,c(recovered.patients,deceased.patients))
#selected.cytokine.profiles <- subset(selected.cytokine.profiles,!(subject_id %in% citeseq.patients)) 
parameter.labels <- unique(selected.cytokine.profiles[,c("test_id","test_name")])
rownames(parameter.labels) <- parameter.labels$test_id
parameter.labels$test_name <- gsub("/ ","/",parameter.labels$test_name,fixed = T)
```


### Grouping of samples
Separate samples into interval groups relative to juncture: early (7-16 days), mid (juncture/17-23 days), and late (24-30 days)
```{r build}
selected.cytokine.profiles$unique.id <- make.unique(apply(selected.cytokine.profiles[,c("subject_id","days_from_admission_to_test","test_id")],1,
                                                          paste0,collapse="."),sep = "|")
selected.cytokine.profiles$unique.id <- sapply(selected.cytokine.profiles$unique.id,function(x){unlist(strsplit(x,"|",fixed=T))[2]})
selected.cytokine.profiles.matrix <- reshape2::dcast(selected.cytokine.profiles,
                                                     subject_id + age + sex + outcome + severity + days_from_symptom_onset_to_test + unique.id ~ test_id,
                                                     value.var = "test_value")
selected.cytokine.profiles.matrix[,colnames(selected.cytokine.profiles.matrix) %in% selected.cytokine.profiles$test_id] <-
  apply(selected.cytokine.profiles.matrix[,colnames(selected.cytokine.profiles.matrix) %in% selected.cytokine.profiles$test_id],2,function(x){scale(log10(x+1))})
selected.cytokine.profiles.matrix$stage <- NA
tmp <- which(selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test >= 7 & selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test <= 16)
selected.cytokine.profiles.matrix[tmp,"stage"] <- "early"
tmp <- which(selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test >= 17 & selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test <= 23)
selected.cytokine.profiles.matrix[tmp,"stage"] <- "mid"
tmp <- which(selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test >= 24 & selected.cytokine.profiles.matrix$days_from_symptom_onset_to_test <= 30)
selected.cytokine.profiles.matrix[tmp,"stage"] <- "late"

cat("Sample distribution:\n")
table(selected.cytokine.profiles.matrix[,c("outcome","stage")])
```
<br/>
    
#### Protein Search 
Search for divergent proteins around juncture using these two models:

Differential expression: 
test_value ~ outcome + age.numeric + sex + DFSO + (1|subject_id), for samples within each stage to identify cytokines
with sig. differences between the two outcome groups

Change between intervals: 
test_value ~ outcome*stage + age.numeric + sex + DFSO + (1|subject_id), the outcome:stage term captures difference
in differences between the two groups entering the juncture and post-juncture
```{r search,out.width=500,fig.height=4}
search.results <- juncture.divergent.proteins(selected.cytokine.profiles.matrix,pval.cutoff = 1)
parameter.labels <- unique(selected.cytokine.profiles[,c("test_id","test_name")])
rownames(parameter.labels) <- parameter.labels$test_id
parameter.labels$test_name <- gsub("/ ","/",parameter.labels$test_name,fixed = T)
search.results$type1$cytokine.label <- parameter.labels[search.results$type1$cytokine,"test_name"]
search.results$type1$de.coef.outcomeDeceased.diff.from.early <- search.results$type1$de.coef.outcomeDeceased - 
                                                                      search.results$type1$de.coef.outcomeDeceased.early
search.results$type1$de.coef.outcomeDeceased.inferred <- search.results$type1$did.coef.outcomeDeceased + 
  search.results$type1$did.coef.test.intervalTRUE.outcomeDeceased


# bubble plots of DE and change effects
#pdf(file.path(output.folder,"critical.juncture","deceased.v.recovered.sig.de.and.change.cytokines.pdf"),width=3,height = 3)

# correlation between different models
mid.did.de.df <- subset(search.results$type1,stage == "mid")
ggplot(mid.did.de.df,aes(did.coef.outcomeDeceased,de.coef.outcomeDeceased.early,label=cytokine.label)) + 
  geom_point(alpha=0.5) + ggrepel::geom_text_repel(data=subset(mid.did.de.df,cytokine.label %in% c("Reg3A","CD25/IL-2R alpha")),size=3) +
  stat_cor() + geom_abline(intercept = 0,slope = 1, color="red", linetype = "dashed") +
  theme_bw()  + labs(title="Pre Juncture",x="Mid v. Early Delta Model ES",y="Early DE Model ES")
ggplot(mid.did.de.df,aes(de.coef.outcomeDeceased.inferred,de.coef.outcomeDeceased,label=cytokine.label)) + 
  geom_point(alpha=0.5) + ggrepel::geom_text_repel(data=subset(mid.did.de.df,cytokine.label %in% c("Reg3A","CD25/IL-2R alpha")),size=3) +
  stat_cor() + geom_abline(intercept = 0,slope = 1, color="red", linetype = "dashed") +
  theme_bw()  + labs(title="Juncture",x="Mid v. Early Delta Model ES",y="Mid DE Model ES")

sig.mid.did.de.df <- subset(mid.did.de.df,de.pval.1 <= 0.05 & did.pval.test.intervalTRUE.outcomeDeceased <= 0.05) # DE in mid & sig. delta 
# order the cytokine by delta ES
sig.mid.did.de.df$cytokine.label <- factor(sig.mid.did.de.df$cytokine.label,
                                           levels=sig.mid.did.de.df$cytokine.label[order(sig.mid.did.de.df$de.coef.outcomeDeceased.diff.from.early)])
ggplot(sig.mid.did.de.df,aes(cytokine.label,de.coef.outcomeDeceased.early)) + 
  geom_hline(yintercept = 0,linetype="dashed") +
  geom_point(aes(size=-log10(de.pval.1.early),color=de.pval.1.early<=0.05,fill="#4bc26c"),alpha=0.75,pch=21) +
  geom_point(aes(y=de.coef.outcomeDeceased,size=-log10(de.pval.1),color=de.pval.1<=0.05,fill="#EADCFA"),alpha=0.75,pch=21) +
  coord_flip() + ylab("Effect Size for Deceased Group") + xlab("Cytokine") + 
  scale_color_manual(name="Sig. P-Value",values=c("white","black")) + scale_fill_manual(name="Stage",labels=c("early","mid"),values=c("#4bc26c","#EADCFA")) +
  scale_size(name="-log10(p-val)",range = c(2,5)) + theme_bw() + theme(text = element_text(size=8),legend.position = "bottom",legend.direction = "vertical")

# early vs. late
late.did.de.df <- subset(search.results$type1,stage == "late")
# correlation between different models
ggplot(late.did.de.df,aes(did.coef.outcomeDeceased,de.coef.outcomeDeceased.early,label=cytokine.label)) + 
  geom_point(alpha=0.5) + ggrepel::geom_text_repel(data=subset(late.did.de.df,cytokine.label %in% c("Reg3A","CD25/IL-2R alpha")),size=3) +  
  stat_cor() + geom_abline(intercept = 0,slope = 1, color="red", linetype = "dashed") +
  theme_bw()  + labs(title="Pre-juncture",x="Late v. Early Delta Model ES",y="Early DE Model ES")
ggplot(late.did.de.df,aes(de.coef.outcomeDeceased.inferred,de.coef.outcomeDeceased,label=cytokine.label)) + 
  geom_point(alpha=0.5) + ggrepel::geom_text_repel(data=subset(mid.did.de.df,cytokine.label %in% c("Reg3A","CD25/IL-2R alpha")),size=3) +
  stat_cor() + geom_abline(intercept = 0,slope = 1, color="red", linetype = "dashed") +
  theme_bw()  + labs(title="Post-juncture",x="Late v. Early Delta Model ES",y="Late DE Model ES")

sig.late.did.de.df <- subset(late.did.de.df,de.pval.1 <= 0.05 & did.pval.test.intervalTRUE.outcomeDeceased <= 0.05) # DE in mid & sig. delta
sig.late.did.de.df$cytokine.label <- factor(sig.late.did.de.df$cytokine.label,
                                        levels=sig.late.did.de.df$cytokine.label[order(sig.late.did.de.df$de.coef.outcomeDeceased.diff.from.early)])
ggplot(sig.late.did.de.df,aes(cytokine.label,de.coef.outcomeDeceased.early)) + 
  geom_hline(yintercept = 0,linetype="dashed") +
  geom_point(aes(size=-log10(de.pval.1.early),color=de.pval.1.early<=0.05,fill="#4bc26c"),alpha=0.75,pch=21) +
  geom_point(aes(y=de.coef.outcomeDeceased,size=-log10(de.pval.1),color=de.pval.1<=0.05,fill="lightblue"),alpha=0.75,pch=21) +
  coord_flip() + ylab("Effect Size for Deceased Group") + xlab("Cytokine") + 
  scale_color_manual(name="Sig. P-Value",values=c("white","black")) + scale_fill_manual(name="Stage",labels=c("early","late"),values=c("#4bc26c","lightblue")) +
  scale_size(name="-log10(p-val)",range = c(2,5)) + theme_bw() + theme(text = element_text(size=8),legend.position = "bottom",legend.direction = "vertical")
#dev.off()

# delta of delta p-value plot
#pdf(file.path(output.folder,"critical.juncture","deceased.v.recovered.sig.de.and.change.cytokines.delta.pval.pdf"),width=1.75,height = 1.85)
# early vs. mid
ggplot(sig.mid.did.de.df,aes(cytokine.label,-log10(did.pval.test.intervalTRUE.outcomeDeceased))) +
  geom_hline(yintercept = -log10(0.05),linetype = "dashed",color="red") + 
  geom_bar(stat="identity",fill="deepskyblue4",alpha=0.75) + 
  scale_y_continuous(breaks = c(0,1,2)) + 
  coord_flip() + ylab("-log10(p-value)") + xlab("Cytokine") + 
  #labs(title="Deceased v. Recovered",subtitle = "At early (7-16 days) and mid (17-23 days) stages") +
  scale_color_manual(name="Sig. P-Value",values=c("white","black")) + scale_fill_manual(name="Stage",values=c("#4bc26c","#EADCFA","lightblue")) +
  scale_size(name="-log10(p-val)",range = c(2,5)) + theme_classic2() + theme(text = element_text(size=8),legend.position = "bottom",legend.direction = "vertical")
ggplot(sig.late.did.de.df,aes(cytokine.label,-log10(did.pval.test.intervalTRUE.outcomeDeceased))) +
  geom_hline(yintercept = -log10(0.05),linetype = "dashed",color="red") + 
  geom_bar(stat="identity",fill="deepskyblue4",alpha=0.75) + 
  scale_y_continuous(breaks = c(0,1,2)) + 
  coord_flip() + ylab("-log10(p-value)") + xlab("Cytokine") + 
  #labs(title="Deceased v. Recovered",subtitle = "At early (7-16 days) and mid (17-23 days) stages") +
  scale_color_manual(name="Sig. P-Value",values=c("white","black")) + scale_fill_manual(name="Stage",values=c("#4bc26c","#EADCFA","lightblue")) +
  scale_size(name="-log10(p-val)",range = c(2,5)) + theme_classic2() + theme(text = element_text(size=8),legend.position = "bottom",legend.direction = "vertical")
#dev.off()
bin1.cytokines <- list("mid"=sig.mid.did.de.df$cytokine,"late"=sig.late.did.de.df$cytokine)
```

```{r correlation,out.width=800,fig.cap="<b>Fig</b> Correlation between outcome and severity comparisons."}
# Correlation with DSM-high and low DE cytokines
dsm.envir <- new.env()
load(file.path(output.folder,"dsm.high.v.low.cytokine.test.results.RData"),verbose = F,envir = dsm.envir)
merged.de.df <- merge(mid.did.de.df[,c("cytokine","cytokine.label","de.coef.outcomeDeceased","de.pval.1")],
                      dsm.envir$divergence.df,by="cytokine",suffixes = c(".deceased.v.recovered","dsm.low.v.high"))
merged.de.df$combined.sig <- paste0(merged.de.df$de.pval.1 <= 0.05,"-",merged.de.df$pval.1 <= 0.05)
ggplot(merged.de.df,aes(de.coef.outcomeDeceased,-coef.dsm.groupDSM.low,label=cytokine.label.deceased.v.recovered)) + 
  geom_point(alpha=0.75,pch=21,size=1,aes(fill=combined.sig)) +
  ggrepel::geom_text_repel(size=1.5,force=3) + geom_hline(yintercept = 0) + geom_vline(xintercept = 0) + 
  theme_bw() + xlab("Deceased vs. Recovered") + ylab("DSM High vs. Low") + scale_fill_uchicago(name="Sig p-value in outcome/severity ")

cat("Concordance of DE results:\n")
cont.table <- data.frame(outcome=sign(merged.de.df$de.coef.outcomeDeceased),severity=-sign(merged.de.df$coef.dsm.groupDSM.low))
table(cont.table)
fisher.test(table(cont.table))
```

### Time course of selected cytokines
```{r time.course,out.width=900,fig.height=5,fig.cap=c("<b>Fig</b> Time course of selected circulating protein levels of critical ill patients with recovery or deceased outcomes. Longitudinal samples from the same individual are connected by gray lines. Trajectories were fitted to the two age- and gender-matched recovered versus deceased patient groups separately and are shown with the shaded areas representing standard error.")}
display.cytokine.groups <- unique(unlist(bin1.cytokines))

display.cytokine.profiles <- subset(selected.cytokine.profiles,test_id %in% display.cytokine.groups)
#display.cytokine.profiles$test_name <- factor(display.cytokine.profiles$test_name,levels=unlist(display.cytokine.groups)) 

display.cytokine.profiles.avg <- aggregate(test_value ~ subject_id + days_from_symptom_onset_to_test + test_name + test_id + outcome,
                                            display.cytokine.profiles,mean)
subjects_with_early_tp <- unique(subset(display.cytokine.profiles.avg,days_from_symptom_onset_to_test <= 23)$subject_id)
ggplot(display.cytokine.profiles.avg,aes(days_from_symptom_onset_to_test,log10(test_value+1),group=outcome)) + 
  #geom_vline(xintercept = 20,color="grey60",linetype="dashed") +
  geom_rect(aes(xmin=17,xmax=23,ymax=Inf,ymin=-Inf),alpha=0.05,fill = "#EADCFA") +
  geom_point(alpha=.25,pch=21,aes(fill=outcome),size=.5) + 
  geom_smooth(se = T,aes(color=outcome,fill=outcome),alpha=0.15,size=.8) + #,data = subset(display.cytokine.profiles.avg,subject_id %in% subjects_with_early_tp)) + 
  geom_line(aes(group=subject_id),alpha=0.1) +
  scale_color_manual(name="Outcome",values=c("#374E55FF","#DF8F44FF")) + scale_fill_manual(name="Outcome",values=c("#374E55FF","#DF8F44FF")) +
  facet_wrap(~test_name,scales = "free",ncol=5,dir = "v") + theme_bw() + xlim(0,60) + 
  xlab("Days from Symptom Onset") + ylab("Cytokine Level (log10 pg/mL)") + 
  theme(strip.background = element_rect(fill="white"),strip.text = element_text(size=7))
```
<br/>

### Serology time course differences
SARS-Covid2 Spike and nucleocapsid antibodies
```{r serology,out.width=600,fig.height=3,fig.cap="<b>Fig</b> Antibody time course for deceased and recovered patients."}
selected.sero.profiles <- subset(covid19.lab.results,test_type == "Serology" & subject_id %in% selected.cytokine.profiles$subject_id)
selected.sero.profiles$test_value <- log10(selected.sero.profiles$test_value+1)

ggplot(selected.sero.profiles,aes(days_from_symptom_onset_to_test,test_value)) +
  geom_vline(xintercept = 20,color="grey60",linetype="dashed") +
  geom_point(alpha=.25,pch=21,aes(fill=outcome),size=1) + 
  facet_wrap(~test_name,scales = "free",ncol = 2) + theme_bw() + 
  geom_smooth(se = T,aes(color=outcome,fill=outcome),alpha=0.15,size=1.25) + 
  geom_line(aes(group=subject_id),alpha=0.1) +
  scale_fill_manual(name="Outcome",values=c("#374E55FF","#DF8F44FF")) + 
  scale_color_manual(name="Outcome",values=c("#374E55FF","#DF8F44FF")) + 
  ylab("Antibody Level (log10 LU)") + xlab("Days from Symptom Onset") 
```

### CBC cell count time course
```{r cell.count,out.width=900,fig.height=3,fig.cap=c("<b>Fig</b> Cell populations as % parent for deceased and recovered patients.","<b>Fig</b> Absolute cell counts for deceased and recovered patients.")}
selected.cbc.profiles <- subset(covid19.lab.results,test_type == "Clinical" & subject_id %in% c(recovered.patients,deceased.patients))
selected.cbc.profiles <- selected.cbc.profiles[grep("((T|C|s|\\.)_x10.[36].uL$)|[sCT]_pct$",selected.cbc.profiles$test_id),]
selected.cbc.profiles <- subset(selected.cbc.profiles,test_id %in% names(which(table(selected.cbc.profiles$test_id) > 100)))

# reference range
cell.counts.range <- unique(selected.cbc.profiles[,c("test_name","test_id","normal_range")])
cell.counts.range[,c("lower.bound","upper.bound")] <- t(sapply(cell.counts.range$normal_range,function(x){as.numeric(unlist(strsplit(x,"-")))}))
cell.counts.range <- cbind(aggregate(lower.bound ~ test_id + test_name,cell.counts.range,min),
                           upper.bound=aggregate(upper.bound ~ test_id + test_name,cell.counts.range,max)[,"upper.bound"])

# remove possible erroneous values
tolerance <- 10
to.remove.rows <- c()
for (i in unique(selected.cbc.profiles$test_id)) {
  cat(i,"\n")
  test.values <- subset(selected.cbc.profiles,test_id == i)$test_value
  median.value <- median(test.values,na.rm = T)
  range <- quantile(test.values,.75,na.rm = T) - quantile(test.values,.25,na.rm = T)
  to.remove.rows <- c(to.remove.rows,rownames(subset(selected.cbc.profiles,test_id == i & 
                                                       (test_value > (median.value + tolerance*range) | 
                                                       test_value < (median.value - tolerance*range)))))
}
#View(selected.cbc.profiles[to.remove.rows,])
selected.cbc.profiles <- subset(selected.cbc.profiles,!(rownames(selected.cbc.profiles) %in% to.remove.rows))

selected.cbc.profiles.avg <- aggregate(test_value ~ subject_id + days_from_symptom_onset_to_test + test_name + test_id + severity.outcome,
                                            selected.cbc.profiles,mean)
to.display <- selected.cbc.profiles.avg[grep("_pct$",selected.cbc.profiles.avg$test_id),]
ggplot(to.display) + 
  geom_hline(data=subset(cell.counts.range,test_id %in% to.display$test_id),aes(yintercept=lower.bound),alpha=1,color="darkseagreen3",linetype="dashed") +   
  geom_hline(data=subset(cell.counts.range,test_id %in% to.display$test_id),aes(yintercept=upper.bound),alpha=1,color="darkseagreen3",linetype="dashed") +
  geom_rect(aes(xmin=17,xmax=23,ymax=Inf,ymin=-Inf),alpha=0.05,fill = "#EADCFA") +
  #geom_vline(xintercept = 20,color="grey60",linetype="dashed") +
  geom_line(aes(x=days_from_symptom_onset_to_test,y=test_value,group=subject_id),alpha=0.1) + 
  geom_point(alpha=.1,pch=21,aes(fill=severity.outcome,x=days_from_symptom_onset_to_test,y=test_value)) +
  geom_smooth(se = T,aes(x=days_from_symptom_onset_to_test,y=test_value,group=severity.outcome,color=severity.outcome),alpha=0.3) + xlim(0,40) +
  scale_color_manual(name="Severity-Outcome",values=severity.color) + scale_fill_manual(name="Severity-Outcome",values=severity.color) +
    facet_wrap(~test_name,scales = "free",ncol=5,dir = "v") + theme_bw() + ylab("Cell Counts (% of Parent Pop.)") + xlab("Days from Symptom Onset") +
    theme(strip.background = element_rect(fill="white"),strip.text = element_text(size=7))

to.display <- selected.cbc.profiles.avg[grep("_pct$",selected.cbc.profiles.avg$test_id,invert = T),]
ggplot(to.display) + 
 geom_hline(data=subset(cell.counts.range,test_id %in% to.display$test_id),aes(yintercept=lower.bound),alpha=1,color="darkseagreen3",linetype="dashed") +   
  geom_hline(data=subset(cell.counts.range,test_id %in% to.display$test_id),aes(yintercept=upper.bound),alpha=1,color="darkseagreen3",linetype="dashed") +
  geom_rect(aes(xmin=17,xmax=23,ymax=Inf,ymin=-Inf),alpha=0.05,fill = "#EADCFA") +
  #geom_vline(xintercept = 20,color="grey60",linetype="dashed") +
  geom_line(aes(x=days_from_symptom_onset_to_test,y=test_value,group=subject_id),alpha=0.1) + 
  geom_point(alpha=.1,pch=21,aes(fill=severity.outcome,x=days_from_symptom_onset_to_test,y=test_value)) +
  geom_smooth(se = T,aes(x=days_from_symptom_onset_to_test,y=test_value,group=severity.outcome,color=severity.outcome),alpha=0.3) + xlim(0,40) +
  scale_color_manual(name="Severity-Outcome",values=severity.color) + scale_fill_manual(name="Severity-Outcome",values=severity.color) +
    facet_wrap(~test_id,scales = "free",ncol=5,dir = "v") + theme_bw() + ylab("Cell Counts") + xlab("Days from Symptom Onset") + 
    theme(strip.background = element_rect(fill="white"),strip.text = element_text(size=7))
```

### Session Info
```{r save}
# save all results
save(merged.de.df,mid.did.de.df,late.did.de.df,file = file.path(output.folder,"deceased.v.recovered.cytokine.test.results.RData"))

sessionInfo()
```
