---
title: "T0 Correlation between Gene Set and Cytokine Expression"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Calculate the correlation of cytokine level and gene set module scores of T0 samples of in the CITESEQ cohort
```{r setup, include=FALSE}
data.folder <- "input.data"
output.folder <- "output/cytokine.geneset.correlation"
dir.create(output.folder,recursive = T)
library(knitr)
library(ggplot2)
library(Biobase)
library(ggsci)
library(pheatmap)
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE)
severity.color <- c("Critical-Alive"="#374E55FF","Critical-Deceased"="#DF8F44FF","Moderate-Alive"="#00A1D5FF" ,"Severe-Alive"="#B24745FF")
```

### Import Data
Use GSVA scores from the PC1_module_score_gsva_filtered_samples_genes file, which was created using edge genes of GSEA of PC1 coefficients.
```{r import.data}
# selected gene sets from Fig 3 to analyse
selected.gene.sets <- read.csv(file.path(data.folder,"pathway_summary.csv"))

# GSEA stats
# gene sets that are sig. in GSEA (or in manual node list) and are of interest
gsea.res <- read.csv(file.path(data.folder,"gene.set.scores.and.gsea.csv"))
gsea.res$cell.type.pathway <- paste0(gsea.res$pathway,"_",gsea.res$celltype)
sig.gsea.res <- subset(gsea.res,PC1continuous_padj <= 0.2)

# gsva module scores
module.score.type <- "PC1_PC1_module_score_gsva_filtered_samples_genes"
#module.score.type <- "healthy_vs_covid_COVID-Healthy_module_score_gsva_filtered_samples_genes"
gsva.scores <- readRDS(file.path(data.folder,paste0(module.score.type,".rds")))
gsva.scores$cell.type.pathway <- paste0(gsva.scores$pathway,"_",gsva.scores$celltype)
# keep only those that are sig. in GSEA
sig.gsva.scores <- subset(gsva.scores,cell.type.pathway %in% sig.gsea.res$cell.type.pathway & pathway %in% selected.gene.sets$Primary_gene_sets)
sig.gsva.scores[,c("subject_id","visit","batch")] <- do.call("rbind",sapply(as.character(sig.gsva.scores$sample),strsplit,"_"))

# metadata
load(file.path(data.folder,"covid19.metadata.paper1.RData"),verbose=T)
dsm.data <- readRDS(file.path(output.folder,"..","DSM","citeseq.patient.end.points.RDS"))
covid19.patient.demo$PC1 <- dsm.data[covid19.patient.demo$subject_id,]$PC1
sig.gsva.scores <- merge(sig.gsva.scores,subset(covid19.samples,material_type == "PBMC")[,c("subject_id","sample_label","visit","days_from_admission_to_sample_drawn")],
                         by=c("subject_id","visit"),all.x=T)


# cytokine data for patients with module scores
cytokine.data <- subset(covid19.lab.results,test_type == "ELISA" & subject_id %in% sig.gsva.scores$subject_id)
cytokine.labels <- as.character(unique(cytokine.data$test_name))
names(cytokine.labels) <- unique(cytokine.data$test_id)
cytokine.data$test_value_log10 <- log10(cytokine.data$test_value + 0.01)
cytokine.data <- reshape2::dcast(cytokine.data,subject_id + days_from_admission_to_test ~ test_id,value.var = "test_value_log10",fun.aggregate = mean)
cytokine.data <- merge(unique(sig.gsva.scores[,c("subject_id","sample_label","visit","days_from_admission_to_sample_drawn")]),cytokine.data,
                       by.x=c("subject_id","days_from_admission_to_sample_drawn"),by.y=c("subject_id","days_from_admission_to_test"),all.x = T)
```

### Correlation
Calculate the correlation between cytokine levels and gene set expression for 28 T0 samples
```{r corr}
module.score.type <- substring(module.score.type,0,3)
# T0 cytokine data
t0.cytokine.data <- subset(cytokine.data,visit == "T0")

# T0 gene set scores
t0.gsva.scores <- reshape2::dcast(subset(sig.gsva.scores,visit=="T0"),subject_id + visit ~ cell.type.pathway,value.var = "module.score")
t0.gsva.scores <- t0.gsva.scores[match(t0.cytokine.data$subject_id,t0.gsva.scores$subject_id),]
    
# spearman correlation
corr.method = "spearman"
cytokine.geneset.cor <- psych::corr.test(t0.cytokine.data[,-c(1:5)],t0.gsva.scores[,-c(1:2)],adjust = "none",method = corr.method)
cytokine.geneset.cor$fdr <- apply(cytokine.geneset.cor$p,2,p.adjust,method="fdr") 

display.corr <- t(cytokine.geneset.cor$r[,unique(which(cytokine.geneset.cor$fdr < 0.05,arr.ind = T)[,2])])
display.sig <- round(t(cytokine.geneset.cor$fdr[,unique(which(cytokine.geneset.cor$fdr < 0.05,arr.ind = T)[,2])]),3)
display.sig[which(display.sig > 0.05)] <- ""
display.sig[which(display.sig != "")] <- "\u00b7"
hm <- pheatmap(display.corr,color = viridis::viridis_pal()(101),display_numbers = display.sig,number_color = "red",fontsize = 7,fontsize_number = 18,
         main=paste0(corr.method,"'s correlation of T0 samples"),border_color = "white",labels_col = cytokine.labels[colnames(display.corr)],
         filename = file.path(output.folder,paste0("cytokine.geneset.correlation.",corr.method,".",module.score.type,".pdf")),
         cellwidth = 7,cellheight = 7)

corr1 <- data.frame(reshape2::melt(cytokine.geneset.cor$n),r=reshape2::melt(cytokine.geneset.cor$r)[,3],
                p=reshape2::melt(cytokine.geneset.cor$p)[,3],fdr=reshape2::melt(cytokine.geneset.cor$fdr)[,3])

# pearson correlation
corr.method = "pearson"
cytokine.geneset.cor <- psych::corr.test(t0.cytokine.data[,-c(1:5)],t0.gsva.scores[,-c(1:2)],adjust = "none",method = corr.method)
cytokine.geneset.cor$fdr <- apply(cytokine.geneset.cor$p,2,p.adjust,method="fdr") 

display.corr <- t(cytokine.geneset.cor$r[,unique(which(cytokine.geneset.cor$fdr < 0.05,arr.ind = T)[,2])])
display.sig <- round(t(cytokine.geneset.cor$fdr[,unique(which(cytokine.geneset.cor$fdr < 0.05,arr.ind = T)[,2])]),3)
display.sig[which(display.sig > 0.05)] <- ""
display.sig[which(display.sig != "")] <- "\u00b7"
hm <- pheatmap(display.corr,color = viridis::viridis_pal()(101),display_numbers = display.sig,number_color = "red",fontsize = 7,fontsize_number = 18,
         main=paste0(corr.method,"'s correlation of T0 samples"),border_color = "white",labels_col = cytokine.labels[colnames(display.corr)],
         filename = file.path(output.folder,paste0("cytokine.geneset.correlation.",corr.method,".",module.score.type,".pdf")),
         cellwidth = 7,cellheight = 7)
dev.off()
corr2 <- data.frame(reshape2::melt(cytokine.geneset.cor$n),r=reshape2::melt(cytokine.geneset.cor$r)[,3],
                p=reshape2::melt(cytokine.geneset.cor$p)[,3],fdr=reshape2::melt(cytokine.geneset.cor$fdr)[,3])

#write.csv(data.frame(cytokine=cytokine.labels[as.character(corr1$Var1)],spearman=corr1,pearson=corr2),row.names = F,
#          file=file.path(output.folder,"cytokine.geneset.correlation",paste0("cytokine.geneset.correlation.",module.score.type,".corr.csv")))
```

### Visualization
Create scatter plots for selected pairs of cytokines and gene sets
```{r scatter.plot}
gene.sets.of.interest <- c("reactome_Oxidative Stress Induced Senescence_pDC","GO_APOPTOTIC_SIGNALING_PATHWAY_pDC","GO_INTRINSIC_APOPTOTIC_SIGNALING_PATHWAY_pDC",
                           "reactome_Regulation of TP53 Activity_pDC","GO_RESPONSE_TO_TYPE_I_INTERFERON_NK_CD16hi","reactome_Fatty acid metabolism_NK_CD16hi")
cytokines.of.interest <- c("IL.15","IFN.alpha2a")
covid19.patient.demo$severity.outcome <- paste0(covid19.patient.demo$severity,"-",covid19.patient.demo$outcome)

for (cytokine in cytokines.of.interest) {
  for (gs in gene.sets.of.interest) {
    df <- data.frame(covid19.patient.demo[t0.cytokine.data$subject_id,],source=t0.cytokine.data[,cytokine],target=t0.gsva.scores[,gs])
    
    print(ggplot(df,aes(PC1,source)) + 
      geom_point(aes(fill=severity.outcome),size=3,pch=21,color="white") + scale_fill_manual(values = severity.color) +
      ggpubr::stat_cor(size=2.5) + theme_bw() + stat_smooth(method="lm",se=F,alpha=0.5,geom="line",color="blue") + xlab("DSM") + ylab(cytokine) + 
      theme(legend.position = "none",text = element_text(size=7)))
    df$source.PC1.resid <- resid(lm(source ~ PC1,df))[rownames(df)] 
    df$target.PC1.resid <- resid(lm(target ~ PC1,df))[rownames(df)] 
    print(ggplot(df,aes(source.PC1.resid,target.PC1.resid)) + 
      geom_point(aes(fill=severity.outcome),size=3,pch=21,color="white") + scale_fill_manual(values = severity.color) +
      ggpubr::stat_cor(size=2.5) + theme_bw() + stat_smooth(method="lm",se=F,alpha=0.5,geom="line",color="blue") + 
      ylab(paste0(gs,".resid")) + xlab(paste0(cytokine,".resid")) + 
      theme(legend.position = "none",text = element_text(size=7)))
  }
}
```

### Additional check
```{r add}
source.gs <- "reactome_Oxidative Stress Induced Senescence_pDC"
target.gs <- "GO_APOPTOTIC_SIGNALING_PATHWAY_pDC"
df <- data.frame(covid19.patient.demo[t0.cytokine.data$subject_id,],source=t0.gsva.scores[,source.gs],target=t0.gsva.scores[,target.gs])
print(ggplot(subset(df,!is.na(PC1)),aes(source,target)) + 
        geom_point(aes(fill=severity.outcome),size=3,pch=21,color="white") + scale_fill_manual(values = severity.color) +
        ggpubr::stat_cor(size=2.5) + theme_bw() + stat_smooth(method="lm",se=F,alpha=0.5,geom="line",color="blue") + xlab(source.gs) + ylab(target.gs) + 
        theme(legend.position = "none",text = element_text(size=7)))

# differences with steroid use
# correlation with glucose response
gene.sets.of.interest <- c("reactome_Fatty acid metabolism_NK_CD16hi","HALLMARK_TNFA_SIGNALING_VIA_NFKB_NK_CD16hi",
                           "HALLMARK_INFLAMMATORY_RESPONSE_NK_CD16hi","HALLMARK_MTORC1_SIGNALING_NK_CD16hi")
sample.steroid.info <- read.csv(file.path(data.folder,"pbmc.samples.csv"))
t0.gsva.scores <- merge(t0.gsva.scores,sample.steroid.info[,c("subject_id","visit","steroid.use")],by=c("subject_id","visit"))
t0.gsva.scores$steroid.use <- t0.gsva.scores$steroid.use != ""
glucResponseGenes_NK_CD16hi <- readRDS(file.path(data.folder,"glucResponseGenesforModel_module_score_gsva_filtered_samples_genes.rds"))
glucResponseGenes_NK_CD16hi <- subset(glucResponseGenes_NK_CD16hi,celltype == "NK_CD16hi")
glucResponseGenes_NK_CD16hi <- glucResponseGenes_NK_CD16hi[grep("T0",glucResponseGenes_NK_CD16hi$sample),]
glucResponseGenes_NK_CD16hi$subject_id <- sapply(as.character(glucResponseGenes_NK_CD16hi$sample),function(x){unlist(strsplit(x,"_"))[[1]]})
t0.gsva.scores.melted <- reshape2::melt(cbind(t0.gsva.scores,subject_id=rownames(t0.gsva.scores)),measure.vars=gene.sets.of.interest)
t0.gsva.scores.melted$variable <- gsub("_NK_CD16hi","",t0.gsva.scores.melted$variable)
t0.gsva.scores.melted$glucResponseGenes_NK_CD16hi <- glucResponseGenes_NK_CD16hi[match(t0.gsva.scores.melted$subject_id,
                                                                                       glucResponseGenes_NK_CD16hi$subject_id),"module.score"]
t0.gsva.scores.melted <- data.frame(covid19.patient.demo[t0.gsva.scores.melted$subject_id,],
                                    t0.gsva.scores.melted[,c("steroid.use","variable","value","glucResponseGenes_NK_CD16hi")])
#pdf(file.path(output.folder,"NK_steroid_response.pdf"),width=4,height=4)
ggplot(subset(t0.gsva.scores.melted,!is.na(PC1)),aes(glucResponseGenes_NK_CD16hi,value,group=steroid.use)) + 
  geom_point(aes(fill=severity.outcome,shape=steroid.use),size=2,color="white") + scale_fill_manual(values = severity.color) +
  ggpubr::stat_cor(size=2.5) + theme_bw() + stat_smooth(method="lm",se=F,alpha=0.5,size=1,geom="line",aes(color=steroid.use)) + 
  scale_shape_manual(values=c(22,21)) + scale_color_d3() +
  theme(legend.position = "none",text = element_text(size=7)) + facet_wrap(.~variable) + ylab("Module Score")
#dev.off()

# correlation with steroid-linked gene - TSC22d3
tsc22d3.expr <- readRDS(file.path(data.folder,"NK_CD16hi_TSC22D3.exp.RDS"))
t0.tsc22d3.expr <- exprs(tsc22d3.expr[,grep("T0",colnames(tsc22d3.expr))])
colnames(t0.tsc22d3.expr) <- sapply(colnames(t0.tsc22d3.expr),function(x){unlist(strsplit(x,"_"))[[1]]})
t0.gsva.scores.melted <- reshape2::melt(cbind(t0.gsva.scores,subject_id=rownames(t0.gsva.scores)),measure.vars=gene.sets.of.interest)
t0.gsva.scores.melted$TSC22D3.expr <- t0.tsc22d3.expr[,match(t0.gsva.scores.melted$subject_id,colnames(t0.tsc22d3.expr))]
t0.gsva.scores.melted <- data.frame(covid19.patient.demo[t0.gsva.scores.melted$subject_id,],t0.gsva.scores.melted[,c("steroid.use","variable","value","TSC22D3.expr")])
#pdf(file.path(output.folder,"NK_TSC22D3_corr.pdf"),width=4,height=4)
t0.gsva.scores.melted$variable <- gsub("_NK_CD16hi","",t0.gsva.scores.melted$variable)
ggplot(subset(t0.gsva.scores.melted,!is.na(PC1)),aes(TSC22D3.expr,value,group=steroid.use)) +
  geom_point(aes(fill=severity.outcome,shape=steroid.use),size=2,color="white") + scale_fill_manual(values = severity.color) +
  ggpubr::stat_cor(size=2.5) + theme_bw() + stat_smooth(method="lm",se=F,alpha=0.5,geom="line",aes(color=steroid.use)) +
  scale_shape_manual(values=c(22,21)) + scale_color_d3() +
  theme(legend.position = "none",text = element_text(size=7)) + facet_wrap(.~variable) + ylab("Scores") + xlab("TSC22D3 mRNA")
#dev.off()
```

### Session Info
```{r save}
sessionInfo()
```
